---
output:
  pdf_document: default
  html_document: default
---
# Data stuctures {#datastructures}

This chapter lists the most common objects to store data. To make the best of the R language, you need a strong understanding of the basic data types and data structures and how to operate on those.

It is very important to understand because these are the objects to manipulate on a day-to-day basis in R. Dealing with object conversions is one of the most common sources of frustration for beginners.

To understand computations in R, two slogans are helpful:

Everything that exists is an object.
Everything that happens is a function call.


## Atomic vectors
The basic data structure in R is the vector. Vectors have three characteristics:

  - a type, `typeof()`,
  - a length, `length()`,
  - some attributes, `attributes()`. 
  
To illustrate:

typeof()  # what is it?
length()  # how long is it? What about two dimensional objects?
attributes()  # does it have any metadata?

e.g.:

```{r}
x <- "dataset"
typeof(x)
attributes(x)

y <- 1:10
typeof(y)
length(y)
attributes(y)

z <- c(1L, 2L, 3L)
typeof(z)
```

### Types of data 

There are five common types of atomic vectors:

  - logical,
  - complex
  - integer,
  - double (often called numeric- real or decimal),
  - character.

Note that if vector elements not all of the same type, R makes coercion.  In that case, the order becomes:
"logical < numeric < character".

Here are a few illustrations.


```{r}
# logical vector
log_vector <- c(TRUE, FALSE, FALSE)
log_vector

# numeric vector
num_vector <- c(12, 10, 3)
typeof(num_vector)

# integer vector
int_vector <- c(12L, 10L, 3L)
typeof(int_vector)


# character vector
chr_vector <- c("a", "b", "c")
typeof(chr_vector)

# mixed types vector
mix_vector <- c(2, "a")
is.numeric(mix_vector)
typeof(mix_vector)
```

Numeric vectors can be created with the shortcut `:`, which is used to imply all the integers from the number on the left of `:` to the number on its right. Even if it is not good practice, these do not require the `c()` call. 

```{r vectors}
vector_A <- 1:5
vector_A
length(vector_A)
```

### Factor vector

This is a special type of vector. It has a limited number of values, called levels. These levels can be unordered (e.g., gender is either "Female" or "Male") or ordered (e.g. school level is "Primary", "Secondary", "Tertiary")
```{r factor}
gender <- factor(c("Male", "Male", "Female", "Male", "Female", "Female", "Male"))
gender
levels(gender)
summary(gender)
school <- factor(c("Primary", "Secondary", "Tertiary"),
					ordered=TRUE)
school

school2 <- factor(c("Primary", "Secondary","Secondary", "Tertiary"),
					labels=c( "Secondary", "Tertiary","Primary"),
					ordered=TRUE)
school2

```

## Matrices and arrays

Matrices are a special vector in R. They are not a separate type of object but simply an atomic vector with dimensions added on to it. Matrices have rows and columns. However R is not often used for matrices calculations: it is too slow for that and there are better programs for it out there (e.g. Matlab).

```{r}
# populate a matrix with the elements in of the vector, and give the dimensions
M <- matrix(c(4, 1, 0, 3, 6, 8), nrow=3, ncol=2) 
M
```

If we think of a matrix as a 2 dimensions vector, then arrays are $n$ dimensions vectors. Is it important? It might be in some very specific cases.

```{r array example}
mya <- array(data=1:18, dim=c(2,3,3))
mya
```

## Lists
These are the one-size fit all structure... A list is an object composed of any other object, even... another list! Very useful data structure!

```{r list example}
school<-factor(c("Primary", "Secondary", "Tertiary"), ordered=TRUE)
mylist <- list(numbers=c(1:60),
				somenames=c("Jim","Jules"),
				results= c(T,F,F,T),
				school=school)
mylist

# change the names of the elements of the list
names(mylist) <- c("N", "O","R","S")
```

## Data frames

Data frames are the second most important data structure in R. We can think of it as a better version of a data set in Excel. It stacks together observations over many variables, each of these variables being a vector. A data frame is a special type of list where every element of the list has same length.

Data frames can have additional attributes such as rownames(), which can be useful for annotating data, like subject_id or sample_id. But most of the time they are not used.

Some additional information on data frames:

Usually created by read.csv() and read.table().
Can convert to matrix with data.matrix()
Coercion will be forced and not always what you expect.
Can also create with data.frame() function.
Find the number of rows and columns with nrow(df) and ncol(df), respectively.
Rownames are usually 1..n.

e.g.:

```{r dataframe}
# mtcars is a built-in data set
data(mtcars)
class(mtcars)
mtcars
head(mtcars)
str(mtcars)

# names of the variables in the data frame
names(mtcars) 
length(mtcars)
nrow(mtcars)
```

### Data frame creation

A data frame can be created manually by providing the function `data.frame` with the corresponding vectors or with the use of one method above. The vectors should be of same length, otherwise R recycles values. The data frame requires a name for each vector.


```{r}
df <- data.frame(let=LETTERS[1:7],
                 num1=10:16, 
                 num2=floor(rnorm(7,100,10)))
df
names(df)
```

```{r}
students <- data.frame(name= c("tr", "ga", "mi", "st", "eb", "ha", "fo", "fi"),
                       age=c(23, 23, 24, 33, 28, 24, 33, 41),
                       like= c(T, T, T, T, F, F, T, T)) 
students
```

Useful functions:

head() - see first 6 rows
tail() - see last 6 rows
dim() - see dimensions
nrow() - number of rows
ncol() - number of columns
str() - structure of each column
names() - will list the names attribute for a data frame (or any object really), which gives the column names.

### Data frame combination

Example for data set combination:

```{r}
df <- data.frame(id = letters[1:10], x = 1:10, y = rnorm(10))
df
```
     id x       y

```{r}
cbind(df, data.frame(z = 4))
```

## A word about attributes

Attributes are important characteristics of vectors (and other data structures).   
They can be seen as some meta data that defines the nature of the object. This is to say that a given attribute can determine how a function applies to an object. Here are three important attributes:

  - the names, `names()`,
  - the dimensions, `dim()`,
  - the class, `class`.

To illustrate the use of attributes, recall the case of a factor vector. The fact that the vector on which it builds is given the attribute of a class (factor) turns that vector into a special one.

```{r}
vec <- c(1.7, 1.3, 4, 3.3, 3.3, 2, 2.3, 2.3)
vec
table(vec)

vec.f <- factor(vec,
                levels=c(1, 1.3, 1.7, 2, 2.3, 2.7, 3, 3.3, 3.7, 4, 5),
                ordered=TRUE)
vec.f
table(vec.f)
```


## Environments

An environment is similar to a bag/list of names. It can be seen as the space were these names "live".   
Environments can include environments. In that case, the enclosing environment is called the parent environment.  
The most common environment is the `globalenv()` also called the `workspace`. 