---
output:
  pdf_document: default
  html_document: default
---
# Getting started {#getset}



## Install applications

Download the following free applications, available on all platforms:

  1. [R](https://cran.uni-muenster.de/) 
  2. [RStudio, free Desktop version](https://www.rstudio.com/products/rstudio/download/#download)
  3. [A Latex distribution](https://www.latex-project.org/get/) (e.g., MacTex for Mac or MiKTeX for Windows machines) 

The first two are easily and quickly installed. The last is a very large program (a few Gb) and needs time to install.  

Another application needed is a 

  4. [Git distribution](https://git-scm.com/downloads)

This is also a free software.  
Once you have installed Git for version control, activate it in RStudio: _Tools> Global Options> Git/SVN_ and click on _Enable version control interface for RStudio projects_.  
Also generate a SHH RSA key. We will use it to identify at the GitHub repo.

## Sign for GitHub

Create an account at GitHub at [https://github.com](https://github.com/join).



## Create your project

_File> New project> Existing Directory_ and chose that book folder.  
Now, every time you create content for your book, you must start a Rstudio session _File> Open Project..._  
All the files of the project are the files of the folder, and vice versa.



## Create GitHub repo and link your machine to it 

Create a new repository (pronounced 'repo') whose name is **exactly** the same as your R project / book folder (e.g., 'myRbook').   
On the top left menu in GitHub, go to _Settings> SHH and GPG keys_ and click on 'New SHH key'. Paste the SHH key generated by RStudio.

In RStudio go to _Tools> Project Options...> Git/SVN_. Under _Version control system_, select 'Git'.    
Still in RStudio, _Tools> Terminal> New Terminal_. This open a Terminal where you type commands. 

Start by initializing Git on the terminal.
```{r, eval=FALSE}
git init
```

Then paste the message shown at the creation of the repo (changing the names, of course):

```{r, eval=FALSE}
git remote add origin https://github.com/YOURNAME/YOURREPO.git
git push -u origin master
```

Your local `master` should now be connected to the `master` on GitHub.  
If necessary, restart RStudio. At the restart, a Git thumbnail should appear in a pane. You are ready to commit and push your files.


### Troubleshooting

If the procedure above works, great! One must be aware, however, that it sometimes hits some problem unexpected and difficult to understand problem. In these cases, a classic solution is to copy and google the error message with a first term such as "RStudio".  
In my experience, the following hack worked. Find the invisible file of the repo _.git> config_. Hidden files on Mac are turned visible by the keyboard key `CMD + SHIFT + .`. In that file, delete the following part (including the space it takes) 

```{r, eval=FALSE}
[remote "origin"]
	url = git@github.com:USERNAME/repo.git
	fetch = +refs/heads/*:refs/remotes/origin/*
```

Then, try again the procedure above.


## Collaboration on GitHub

There are various established ways for collaborating on a GitHub repo. As a matter of facts, collaboration on a project is the _raison d'etre_ of GitHub.  
We illustrate here the easiest of them, namely the joint reading and writing into one repo by all the members of a team.   
Importantly, it is assumed that all members of the group are able to push/pull code from RStudio to an experimental repo on GitHub (see above).   
The following are the steps in the collaboration's setup.  

### Owner of the organization

One member of the group creates an organization on GitHub under _Settings> Organizations> New organization_.  
Then, under that organization, this member of the team creates a new repo whose name is the same as the folder and R project that the group will work on (e.g., 'ouRbook').  
The owner of the organization connects from RStudio to the repo and populates it with the current files of the project.  
Still as a task of the owner, create a team. Under the organization page in GitHub, simply click on 'New team' and give it a name.  
A team now appears in the organization page. In the page of that team, click on 'Add a member'. A window appears where the member to be added can be searched for and added.  
Once that member is found and selected, make the double step of adding him/her to the team AND assign him/her to the repo. This second step can still be made later, but it's simpler to do it right then.   
Importantly, make sure that the new member has reading and **writing** rights into the repo. This again, can be changed later by navigating the team's page.

### Member of the team

After the steps above, the newly added member receives an email to confirm the participation the organization / team.   
Navigate now to the team's page and locate the repo that the owner has created.  
A button-menu allows to 'Clone or download the repo'. Click to show the https address of the repo, `https://github.com/ORGANIZATION/repo.git` and copy that link.  
Open a simple RStudio session (not a project!). Open a new Terminal (_Tools> Terminal> New Terminal_) and type `cd` (change directory) and a path to the folder where you want the repo to be saved. For instance,

```{r, eval=FALSE}
cd Dropbox/Fresenius/DS4B 
```

The terminal is writing into that folder, as indicated by the path before the `$` sign in command line of the terminal.  
Recall that you copied the address of the team's repo. Then, in command line, now type the following commands and your copied address:

```{r, eval=FALSE}
git clone https://github.com/ORGANIZATION/repo.git 
```

The whole repo is now a new folder in your directory. If all went well, you can now pull and push into that repo in GitHub.














