# The first steps {#gettingstarted}


## Installing the required applications

The first step of working with R is to download the following free applications, which are available on the following platforms:

  1. Download and install [R](https://cran.uni-muenster.de/) 
  2. Download and install [RStudio, free Desktop version](https://www.rstudio.com/products/rstudio/download/#download)
  3. Download and install [A Latex distribution](https://www.latex-project.org/get/) (e.g., MacTex for Mac or MiKTeX for Windows machines) 

The first two applications are easily and quickly installed for Mac and Windows. The third application is very large (a few Gb) and needs some time to be installed.  

Another application needed is a 

  4. [Git distribution](https://git-scm.com/downloads)

This application is also classified as a free software.  
Once you have installed Git for version control, activate it in RStudio: _Tools> Global Options> Git/SVN_ and click on _Enable version control interface for RStudio projects_.  
Also generate a SHH RSA key. Generating an SHH RSA key is required to link RStudio to a repository on GitHub.

(GitHub is a website and cloud-based service that helps developers store and manage their code, as well as track and control changes to their code)

## Signing up for GitHub

Create an account at GitHub at [https://github.com](https://github.com/join).



## Creating a new project

RStudio projects are associated with working directories in R. RStudio projects can be created: 

- In a new directory
- In an existing directory (R code and data already exists)
- By cloning a version control repository (Available on Git or any other subversion)

Focusing on our our project in Data Science, we will focus on creating an an RStudio project in an existing directory

_File> New project> Existing Directory_ and chose the suggested book folder. You can also change the location of the project working directory. 

Important: Every time you create new content for your book, you must start an Rstudio session _File> Open Project..._  
All the files of the project are the files of the folder (working directory), and vice versa.



## Create a new GitHub repository 

The next step of the process is to create a new repository (pronounced 'repo') on GitHub.com. 
1. Name the repository **exactly** by the name of the created R project / book folder (e.g., 'myRbook').   
2. Enter a description for the repository (Optional)
3. Select public visibility.
4. Select initialize this repository with a README (Optional)
5. Select add .ignore and select R.
6. Create the repository.

After having created the repository, go to the settings of the respective repo and deploy a new SSH key, which has been generated by RStudio. Go to _Settings> Deploy keys_ and click on 'add deploy key'. Copy and paste the SHH key generated by RStudio: _Tools> Global Options> Git/SVN_ and click on view public key.

In RStudio go to _Tools> Project Options...> Git/SVN_. Under _Version control system_, select 'Git'.    

## Link the repository to RStudio 

Still in RStudio, go to _Tools> Terminal> New Terminal_. This will open a Terminal where commands can be typed in.

Start by initializing Git on the terminal.
```{r, eval=FALSE}
git init
```

Then paste the message shown at the creation of the repo (changing the names, of course):

```{r, eval=FALSE}
git remote add origin https://github.com/YOURNAME/YOURREPO.git
git push -u origin master
```

After finalizing the above described process, the local `master` should now be connected to the `master` on GitHub.  
If necessary, restart RStudio. At the restart, a Git thumbnail should appear in a pane. You are now ready to commit and push your files.

## Commit and push any changes to GitHub

In the next chapter of the book, R Markdown documents or files are explained. In general, a R Markdown file is a record of your research. Whenever RMD files are changed or updated, it is time to commit and push them to your GitHub repository.

1. In RStudio click the "Git" tab in the upper right pane.
2. Click Commit.
3. In the Review changes view, check the staged box for all files you want to commit and push to GitHub.
4. Make sure to always include or add a commit message (e.g. Intial Commit).
5. Click Commit.
6. Click the Pull button to fetch any remote changes from GitHub.
7. Click the Push button to push any changes to the repository on GitHub.
8. On GitHub, navigate to the Code tab of the repository and refresh the page to see the changes.

Important: Before you push any changes to GitHub, make sure to commit the files you want to add.



### Troubleshooting

If this procedure works, great! You're good to go.

Yet: Please be aware that the above described process sometimes hits unexpected problems. These problems are often difficult to understand (especially as a "beginner" of R). If this is the cas, the best sultion is to copy and google the error message. 

The solutions offered on Google will definitely help solve the experienced problem.


## Collaboration on GitHub

There are various established ways for collaborating on a GitHub repo. As a matter of facts, collaboration on a project is the _raison d'etre_ of GitHub.  
We illustrate here the easiest of them, namely the joint reading and writing into one repo by all the members of a team.   
Importantly, it is assumed that all members of the group are able to push/pull code from RStudio to an experimental repo on GitHub (see above).   
The following are the steps in the collaboration's setup.  

### Creation of an organization

The following section focuses on the creation of an organization, which enables a group of people to work on the same project and therefore collaborate via GitHub. Every member of the organization will then have the opportunity to commit, push and pull any changes of the prokject.

Step 1: One member of the group creates an organization on GitHub under _Settings> Organizations> New organization_.  

Step 2: Under the newly created organization, this member of the team creates a new repo whose name is the same as the folder and R project that the group will work on (e.g., 'ouRbook').  

Step 3: The owner of the organization connects RStudio to the repo (Described process) and populates it with the current files of the project.  

### Creation of a team

Step 1: Still as a task of the owner, create a team. Under the organization page in GitHub, simply click on 'New team' and give it a name. A team now appears in the organization page.  In the page of that team, click on 'Add a member'. A window appears where the member to be added can be searched for and added.  

Step 2: Once that member is found and selected, make the double step of adding him/her to the team AND assign him/her to the repo. This second step can still be made later, but it is easier to do it right then.   

Important: Make sure that the new member has reading and **writing** rights to the repo, in order to be able to also make changes. This again can be changed later by navigating the team's page.

### Member of the team

After the steps above, the newly added member receives an email to confirm the participation of the organization and team.   
After accepting or confirming the participation, navigate to the page of the team and locate the repo that the owner has created.  
A button-menu allows to 'Clone or download the repo'. Click to show the https address of the repo, `https://github.com/ORGANIZATION/repo.git` and copy that link.  
Open a simple RStudio session (not a project!). Open a new Terminal (_Tools> Terminal> New Terminal_) and type `cd` (change directory) and a path to the folder where you want the repo to be saved. For instance,

```{r, eval=FALSE}
cd Desktop/DataScience/OurRProject 
```

The terminal is writing into that folder, as indicated by the path before the `$` sign in the command line of the terminal.  
Recall that you copied the address of the team's repo. Then, in the command line, type the following commands and your copied address:

```{r, eval=FALSE}
git clone https://github.com/ORGANIZATION/repo.git 
```

The whole repo is now a new folder in your directory. If all went well, you can now pull and push into that repo in GitHub.














