# `tibble` {#tibble}

A tibble is the equivalent of a data frame in the `tidyverse`. The specific package of the bundle is called `tibble`.  
All the packages in the `tidyverse` work with tibbles.   
Notice that all the functions applicable to a data frame will also work on a tibble. This is because **all tibbles are data frames** (while the reverse is not true, all data frames are not necessarily tibbles).  
So why bother?  
The reasons to choose tibbles include: 

- faster calculations,
- better printing of the content,
- more information about the content,
- more useful warnings of errors,
- simpler and more limited functionalities.

To install the CRAN version:

```{r
install.packages("tibble")
```

## Creating a tibble

There are various ways to create a tibble.

### `as_tibble`

The simplest way is declare a data frame a tibble with the function `as_tibble`. 

```{r}
require(tidyverse)
data(iris)
# print the data frame
iris
t_iris <- as_tibble(iris)
# print the tibble
class(t_iris)
t_iris
```

### Manually

A tibble can also be created manually. Notice the following two points:

- vectors of length 1 (only) are recycled,  
- vectors that depend on others can be directly created.
  
```{r}
example_tibble <- tibble(
  value1 = 1:10, 
  value2 = 5, 
  value3 = value1^2 * value2
)
example_tibble 
```

Another example for manually creating a tibble:

```{r}
tibble(
  x = 1:10, 
  y = x ^ 3, 
  xy = x+y
)
```


### `tribble`

A tribble is a tibble created in a transposed way. To be noticed:

- column headings are entered (as a formula) with a `~`,
- data are entered in rows,
- all values are separated by commas.

```{r}
example_tribble <- tribble(
	~Data1, ~Data2, ~Data3,
	"X", 2, 7,
	"Y", 5, 12,
	"Z", 4, 8 
	)
example_tribble
```