# `readr`  and `readxl`  {#readr}  
 
 


## `readr`



As shown in Chapter \@ref(import), base R has many functions to read some types of files. Readr therefore provides a fast and simple way to read rectangular data, such as csv, fwf etc. 



`readr` improves on them because it:  

- is much faster,
- has mainly functionalities,
- converts data directly into tibbles,
- handles the conversion of the type of data in a better way.


Notice that `readr` uses names of functions that are very close to the base R functions: e.g., `read_csv` instead of `read.csv` in base R.    
The functions form  a `read_xxx` family where 'xxx' stands for the way the data you want read was recorded:

read_csv(): comma separated (CSV) files
read_tsv(): tab separated files
read_delim(): general delimited files
read_fwf(): fixed width files
read_table(): tabular files where columns are separated by white-space.
read_log(): web log files

In my experience, using the right member of the family yields better results, for instance in parsing the columns.

To install the CRAN version:

```{r
install.packages("readr")
```

### `read.csv`

read_csv() reads comma delimited files and is one of the most common forms of data storage. Therefore, if you understand how read.csv() works, the other functions of readr are also fairly easy to understand. 

When you run read_csv() it prints out a column specification that gives the name and type of each column. Example:

```{r}
read_csv
```

The first step of the process is then the selection of the path to the file to read (As shown in the following example CSV File:

```{r}
read_csv("Insurance_sample.csv")
```

The data might not have column names. You can use col_names = FALSE to tell read_csv() not to treat the first row as headings, and instead label them sequentially from X1 to Xn:

```{r
read_csv("Insurance_sample.csv", col_names = FALSE)
```

The insurance example above has column names and it is therefore not necessary to tell read_csv() to treat the first row as headings.

At this point in time, this is all you need to now about reading or treating CSV files. There are many other functions with regards to reading and writing CSV files, yet for the sake of this project, it is not necessary to have that level of detail.

### `read_delim`  

We'll also describe `read_delim` because functions of the family behave are special cases of this function. For instance, `read_csv` is equivalent to `read_delim` with a `,` as a character to separate the columns.  

- The `file` argument gives the name of a file, possibly with the path to it, if it is not in the same folder. Again, the path can also be an url.  
- `delim` provides the character that separates columns in the data.


```{r}
# assuming that the AH001.txt is in the folder
df <- read_delim ("AH001.txt", delim = "," , trim_ws=TRUE)
df
class(df)
```

Notice an important message, the types of the columns were guessed by the function. For instance, the column "`r names(df)[1]`" was guessed to be of type "`r class(names(df)[1])`". With that respect, the next arguments are particularly useful.    
- First, `trim_ws` is a logical for whether the leading white space in each recording should be erased before parsing. With a white space, columns of numeric values with different number of digits could be guessed as character vectors.  
Second, we can override the guess of `read_delim` by specifying the types: this is called "to parse" a file.  
We can parse into many types and use abbreviations to set them: c = character, i = integer, n = number, d = double, l = logical, D = date, T = date time, t = time, ? = guess.

All of these parsing calls can cause unexpected problems. For instance, one may need to change the symbol for decimal with the extra argument for the column type `locale=locale(decimal_mark=",")`. 
Notice the behavior when an observation is not of the same type as expected: it is changed to `NA`.  
The function `problems` lists the rows with problems.

```{r}
df <- read_delim("AH001.txt", delim=",", trim_ws=TRUE,
                col_types = cols(LPrice="d", HPrice ="d") )
df
problems(df)
```

Notice that the names of the variables are not within `" "`. This is a general feature of the `tidyverse`.   
By default, the first line of the data file is used for naming the columns. This can be changed.  
- `col_names` is the vector of names set for the variables.
- `skip` determines how many lines to be skipped in the file. This is useful when the file does contain names that one does not want to keep.

```{r}
df <- read_delim("AH001.txt", delim = "," , trim_ws=TRUE,
                skip = 1, col_names = FALSE)
df
colnames(df) <- LETTERS[1:length(df)]
df
```

Data often contain specific special character, in particular for comments and NA's.

- `comment` takes the character that signals comments.
- `na` signals what values should be consider as NA's.


```r
df <- read_delim("AH001.txt", delim = "," , trim_ws=TRUE,
                skip = 1, col_names = FALSE, 
                comment = "%", na = "9999")
```



## `readxl`

Lots of rectangular data out there is in Excel files. One way is to handle it could be to first transform the data into a `.csv` file and then import it as described above.  
Thanks to the `readxl` package, one can also read it directly.[^Notice that R can also write to an Excel file. Why would one do that? Maybe because collaborators only work with Excel. This is achieved with the package `XLConnect` but we won't dig into these details here.]
As part of the `tidyverse` bundle, `readxl` shares features and rules with `readr`.   
Below are some of the specific features.  

- `read_excel` is the function that reads the data in the file and assigns it to an object, as illustrated in the following chunk.

```{r}
df <- readxl::read_excel("data_china.xlsx")
df 
```
Importantly, notice that the exact extension does not matter as `readxl` recognizes both `.xls` and `.xlsx`.  
Notice that if the file contains multiple sheets, these can all be accessed.

- `excel_sheets` is a function that yields the names of the different sheets in a file.  

```{r}
readxl::excel_sheets("data_china.xlsx")
```

To know the specific sheets allows to call them directly by name. Alternatively, this can also be done by a number.

- The `sheet` argument specifies the required sheet either by a name or by a number.
```r
df <- read_excel("data_china.xlsx", sheet="FJ") 
```

A specific range for the data to be read can also be specified. Here are arguments that allow that.  

- `n_max` for the maximum number of line to be read.
- `range` for a specific range, describe in one of the possible ways:
  - extreme cells of rectangular data: `range="B1:D10"`,
  - with `cell_rows` for the required rows: `range=cell_rows(1:10)`,
  - with `cell_cols` for the required columns: `range=cell_cols("B:E")`.












