# `magrittr` {#magrittr}


The `magrittr` package offers an operator, the pipe `%>%`, which ultimately will improve coding in R by:

- Structuring sequences of data operations from left-to-right
- Minimizing the need for local variables and function definitions
- Avoiding nested function calls
- Simplifying the process of adding steps anywhere in the sequence of operations

In a few words, that are not as clear as the examples below, the functions calls are now decomposed into steps and these steps are "_piped_" thanks to the operator `%>%`.    

To install the CRAN version:

```{r
install.packages("magrittr")
```

## How to pipe

A basic pipeline is constructed under the following general idea: the call to the left of the pipe is used as an argument (by default the first argument) in the call to the right of the pipe.  
Using the mathematical notation for functions, the following statements give a pretty accurate idea of the procedure.  

Basic piping:
x %>% f is equivalent to f(x)
x %>% f(y) is equivalent to f(x, y)
x %>% f %>% g %>% h is equivalent to h(g(f(x)))

Example for `x %>% f` (This is the most basic rule for piping):

```{r}
BasicPipe <- c(1:5, 10, 12)
mean(BasicPipe)
BasicPipe %>%
	mean()
```

Notice that it is usual to break lines after the pipe `%>%` as it facilitates reading and addition of new steps. Also, the `()` pasted to `mean` is not necessary, but, again, it makes it clear that we pipe a vector into a function.  
Generally, the construction of the code must be re-thought as the calls above are read in reverse ways: `mean(BasicPipe)` reads use the function `mean` on the vector `BasicPipe`, while `BasicPipe %>% mean()` reads take the vector `BasicPipe` and apply the function `mean` to it.



Example for `x %>% f(y)`  (This rule shows how the left-to-pipe element is used as first argument in the right-to-pipe call):

Recall this property as the _first argument rule_.

```{r}
BasicPipe <- c(1:5, 10, 12)
mean(BasicPipe, trim=0.1)

BasicPipe %>%
	mean(trim=0.1)
```



Example for `x %>% f %>% g %>% h` (This rule shows that the pipeline can include many steps):

```{r}
BasicPipe <- c(1:5, 10, 12)
round(sqrt(mean(BasicPipe)),2)

BasicPipe %>%
	mean() %>%
	sqrt() %>%
	round(2)
```


## Using a placeholder

One can actually decide where the left-to-pipe element should be included in the right-to-pipe call. For that, use a dot, `.`, as a placeholder in the right-to-pipe call.

- `x %>% f(y, .)` is equivalent to `f(y, x)`

The `.` indicates where the left-to-pipe element should be used.

```{r}
round(17.23893,3)

3 %>%
	round(17.23893, .)

```


- `x %>% f(y, z = .)` is equivalent to `f(y, z = x)`

Notice that the left-to-pipe element need not be an argument, but can also be the value of an argument.


## Multiple placeholders

The placeholder can be used multiple times. Notice, however, that if the placeholder is not an argument (but simply a value of an argument), then `magrittr` still applies the _first argument rule_, unless the call on the right-to-pipe is enclosed in curly brackets.

- `x %>% f(y = nrow(.), z = ncol(.))` is equivalent to `f(x, y = nrow(x), z = ncol(x))`

- `x %>% {f(y = nrow(.), z = ncol(.))}` is equivalent to `f(y = nrow(x), z = ncol(x))`

```{r}
require(tidyverse)
data(iris)


t_iris <- as_tibble(iris)
t_iris2 <- t_iris[t_iris$Sepal.Length==5, ]
av.sw <- mean(t_iris2$Sepal.Width) 
av.sw

av.sw2 <- iris %>%
            as_tibble() %>%
            .[.$Sepal.Length==5, ] %>%
            {mean(.$Sepal.Width)}
av.sw2  
```