---
title: "StringR"
output: html_document
---

The stringr package focuses on and provides you with a set of functions designed to make working with strings as easy as possible. Stringr focusses on the most important and commonly used string manipulation functions.

Let's start with the installation of StringR:

``{r
install.packages("stringr")
```

Usage
All functions in stringr start with str_ and take a vector of strings as the first argument.

```{r}
x <- c("stadium", "football", "players", "halftime", "referee", "goalkeeper")
str_length(x) 

str_c(x, collapse = ", ")

str_sub(x, 1, 3)
```

There are seven main verbs that work with patterns:

str_detect(x, pattern) tells you if there’s any match to the pattern.

```{r}
str_detect(x, "[a]")
```

str_count(x, pattern) counts the number of patterns.

```{r}
str_count(x, "[a]")
```

str_subset(x, pattern) extracts the matching components.

```{r}
str_subset(x, "[r]")
```

str_locate(x, pattern) gives the position of the match.

```{r}
str_locate(x, "[a]")
```

str_extract(x, pattern) extracts the text of the match.

```{r}
str_extract(x, "[a]")
```

str_match(x, pattern) extracts parts of the match defined by parentheses.

# extract the characters on either side of the vowel
```{r}
str_match(x, "(.)[ae](.)")
```

str_replace(x, pattern, replacement) replaces the matches with new text.

```{r}
str_replace(x, "[aei]", "X")
```

str_split(x, pattern) splits up a string into multiple pieces.
```{r}
str_split(c("a,b", "c,d,e"), ",")
```