# `tidyr` {#tidyr}


Among the many ways of organizing a dataset, the `tidyr` way is a convenient, thought-through and consistent way of thinking of a dataset.  
There are many types of messy data. We introduce here the tools to bring each to a tidy version.  
The general principle for creating tidy data is expressed in these three rules:

- every column is a variable,
- every row is an observation,
- every type of observation in a different table.

The tidying of a data set is mainly obtained with two functions/verbs: `gather` and `spread`.  
Further tidying is obtained with the functions/verbs `separate` and `unite`.

To install the CRAN version:

```{r
install.packages("tidyr")
```

## `gather`

`gather` takes multiple columns, and gathers them into key-value pairs. From a wide data set, we obtain a (vertically) longer data set.  
Let's illustrate this verb with an example.

Consider the following data set that gives the gains of each person in a lottery between Monday and Thursday.

```{r}
flights <- tibble(person= c("Philipp","Corinna","Laurin"),
                  jan= c(2, 5, 1),
                  feb= c(3, 7, 4),
                  mar= c(1, 3, 4),
                  apr= c(1, 2, 5)
                  )
flights
```

This data set has 5 columns. Clearly, this does not correspond to 5 variables. Instead, a tidy data set should only include 3 columns for the variables 'name', 'day' and 'gains'. Each row would then be an observation with one value for each of these variables: e.g., Beth won 7 on Tuesday.  
For the tidying, observe that the appropriate variable 'day' is spread across four columns 'mon', 'tue', 'wed' and 'thu'. These must `gather`ed into one column/variable.

- `tidyr` calls `key` the tidy variable whose values are the names of the messy columns
('day' in the example above),   
- `tidyr` calls `value` the tidy variable whose values are spread over the messy cells
('gains' in the example above).

The call of the function is then:

```r
gather(df, key, value, messy_col1, ..., messy_coln)
```

In the example above, we would obtain the tidy data set in the following way.
```{r}

tidy_flights <- gather(flights,
                        key= "months",
                        value= "flights",
                        jan:apr
                        ) 
tidy_flights
```
Notice how `tidyr` understands the name of the columns when separated by `:`. It simply uses all the columns in the positions from (including) the first column given to the last (including) column given.   
Alternatively, one can obtain the same result by excluding variables to be gathered using the minus symbol `-`.

```r
tidy_lottery <- gather(flights, key="months", value="flights", -person)
```


```{r}
wine <- tribble(
  ~name, ~`2018`, ~`2011`, ~`2020`,
  "mie", 10, 20, 30,
  "eba", 1, 2, 3,
  "str", 14, 24, 5
)

wine

tidy_wine <- gather(wine, key=year, value=glassesofwine, `2018`, `2011`, `2020`)
  
tidy_wine
```


## `spread`

`spread` does the opposite of `gather`. It takes two columns (key & value), and spreads into multiple columns: From a (vertically) long data set, we obtain a wider data set.  
Let's illustrate this verb with an example.


```{r}
students <- tibble(student= rep(c("Moritz","Philipp","Niklas"),2),
                  info= rep(c("hair","age"),3),
                  measure= c("red", 23, "brown", 23, "blond", 23)
                  )
students

students3 <- rbind(students, c("Freddy", "height", 185))
students3

students2 <- tibble(
student= c(rep(c("Moritz","Philipp","Niklas"),2),   "Freddy"),
info= c(rep(c("hair","age"),3),    "height"),
measure= c("red", 23, "brown", 23, "blond", 23,   185)
                  )
students2

tidy_students2 <- spread(students2, key=info, value=measure)
tidy_students2
```

Clearly, here, each single observation (of one person) has information over different rows. These must be `spread` into appropriate variables.  
Recall that the `key` refers to the tidy variable _vs_ messy columns.  
Also, recall that `value` refers to the tidy variable _vs_ messy cells.

```{r}
tidy_students <- spread(students, key=info, value=measure)
tidy_students


```



## `separate` 

`separate` is used on a column to separate its content into various columns.  
The simplest call is generally the following:

```r
separate(df, messy_var, into=c(tidy_var1, tidy_var2))
```  

We now illustrate the use of this verb.

```{r}
tennisplayers <- tibble(info=c("Federer, Swiss","Kohlschreiber, 
                               German","Kyrgios, Australian"))
tennisplayers

tidy_tennisplayers <- separate(tennisplayers,
                          info,
                          into= c("name", "nationality")
                          )
tidy_tennisplayers
```

By default, the separation is made on a character that is neither a number nor a letter. But we can also define the separation character.
```{r}
separate(tennisplayers,
          info,
          into= c("name", "nationality"),
          sep= "r"
          )
```

Furthermore, we can also `separate` based on the number of characters in the values of the variable. Positive integers are used to start from the left while negative integers start from the right.

```{r}
separate(tennisplayers,
          info,
          into=c("name", "nationality"),
          sep=5
          )

separate(tennisplayers,
          info,
          into= c("name", "nationality"),
          sep= -4
          )
```

## `unite`

`unite` is used on columns to gather their content into one variable.  
The simplest call is generally the following:

```r
unite(df, tidy_var, messy_var1, messy_var2, sep="")
```

A main use of this verb is to paste characters from various columns.  
By default, these character values are united with a `_` in between, but the separator can be chosen.


```{r}
mail <- unite(tidy_tennisplayers, mail, name, nationality, sep="#")
mail
```





