---
output:
  pdf_document: default
  html_document: default
---
# `dplyr` {#dplyr}


There are essentially two parts in this package, both dealing with manipulating data set.  
The first gives tools to work on a given data frame, the second to combine data frames. We describe both in turn.  

To install the CRAN version:

```{r
install.packages("dplyr")
```

## Grammar of data manipulation  

`dplyr` introduces a easier and faster way to manipulate datasets and extract information from them.
Hence, all the functions introduced here are redundant with other functions in base R. As for the rest of the `tidyverse`, they just happen to do the job in a clearer way (among other advantages).  
The approach builds on "verbs" (functions) applied successively thanks to the pipe operator `%>%`.

`dplyr` is therefore a system or grammar of data manipulation, providing a set of verbs that help you solve the most common data manipulation challenges:

- `select`
- `mutate`
- `filter`
- `arrange`
- `summarize`

These operations can be made on groups thanks to the function `group_by`.  
A few comments before we describe the verbs are in turn.

- These verbs are better used in tidy data: 
  - each column is a variable,  
  - each row is an observation.

- `select` and `mutate` manipulate/apply on the variables (columns).  

- `filter` and `arrange` manipulate/apply on the observations (rows).

- `summarise` manipulates groups of observations.

- Like in other packages of the `tidyverse`, there is no need for the usual quotes `" "` for the names.  


For the examples below, we will be using the `starwars` data set from the `dplyr`package. So, we can first have a glimpse at it.

```{r}
require(dplyr)
starwars
```

## `select`

The call for this function is of the form:

```r
select(df, var1, ..., varn)
```

where `df` is the original data frame from which one wishes to select data.    
Alternatively, we can write the above code. with the pipe operator

```r
df %>%
  select(var1, ..., varn)
```

Notice that the underlying data frame is not modified. However, the extracted data can be saved in the usual way, i.e., as an object with a name. 

```r
piece_df <- df %>%
              select(var1, ..., varn)
```

Here is an example of a selection of variables.

```{r}
require(MASS)
require(dplyr)
starwars %>% 
  dplyr::select(name, species)
# compare with base R

starwars[,c("name", "species")]
```

There are any ways of selecting variables:

```{r}
## columns/variables 1 to 4
dplyr::select(starwars, 1:4) 
## works also with the name
dplyr::select(starwars, name:hair_color) 
## with a negative number for all but some variable
dplyr::select(starwars, -3) 
## works also with the name
dplyr::select(starwars, -mass) 
```


### Helper arguments

These are arguments to help `select` the variables/columns. Here are a few of them (see `?select` for a complete list):

- `starts_with("x")`
- `ends_with("x")`
- `contains("x")`
- `matches("x")`

Notice the use of the quotes `" "` for the strings because they are not variables names.   
For instance, here, we select all the variables that have `color` in their name.
```{r}
require(tidyverse)
starwars %>%
  dplyr::select(
    name,
    contains("color")
    )

starwars %>%
  dplyr::select(name, hair_color, films)
```

## `mutate`

The `mutate` function allows to create new variables/columns, based on the existing variables.   
The generic call is as follows.

```r
mutate(df, new_variable = expression)

# with pipe

df %>%
  mutate(new_variable = expression)
``` 

We play a bit with it.  

```{r}
## add variable bmi
starwars %>%
         mutate(bmi = mass / (height/100)^2 )  

## two variables created, including one created in the same call
df_b <- starwars %>%
          mutate(
            bmi = mass / (height/100)^2,
            sr.bmi = sqrt(bmi)
            ) 
df_b

starwars %>%
  mutate(
    g_b = sample(c("G", "B"),
                 87,
                 replace = TRUE
                 )
    )
```


## `filter`

The `filter` function is equivalent of a selection, but for rows. The call is generically

```r
filter(df, condition)

# with the pipe

df %>%
  filter(condition) 
```

At this stage, it might be useful to recall the operators for conditions under `?Comparison`.  
Here are examples for this function.

```{r}
starwars %>%
  filter(species != "Droid")

starwars %>%
  filter(
    species != "Droid" &
    eye_color == "blu"
    )

# in base R

starwars[starwars$species!="Droid" & starwars$eye_color=="blue" , ]
```


## `arrange`

This function reorders the rows of the data set, by default in ascending order of a given variable. The typical call, for a reordering over `var1`, is:

```r
arrange(df, var1)

# with pipe

df %>%
  arrange(var1)

```

The use of multiple variables for reordering is sometimes necessary in order to break ties.

```r
## in case of several rows with same value of var1, use var2
df %>%
  arrange(
    var1,
    var2)
```

Use `desc` on the variable to show in descending order.  
Here is an example that combines these two points.

```{r, results='hide'}
starwars %>%
        arrange(
          height, 
          desc(mass))

starwars %>%
  arrange(desc(name))
```

## `summarise`

Contrary to the other verbs, the `summarise` function creates a new data frame. The usual call is:

```r
summarise(df, name = expr)

# with the pipe
df %>%
  summarise(name = expr)
```  

where `expr` stands for any function on a vector (most of the time, one of the variables) that returns a single value. That function can either be built-in or a function provided by the user.  
In that sense, notice that `summarise` does not mean to make a summary but, instead, to collapse a full vector into one single value.

```{r}
starwars %>%
  summarise(min_h = min(height, na.rm = TRUE),
            av_m= mean(mass, na.rm = TRUE)
            )
```

### Helper functions

These are functions to help `summarise`, but are essentially wrappers for the function `[[`. These include `first(x)`, `last(x)`, `nth(x,n)`, `n()`... where `x` is the variable.

```{r}
starwars %>%
  summarise(nth(species, 3))
```
Other functions can be used as logical tests.

```{r}
starwars %>%
  summarise(sum(species == "Human", na.rm=TRUE))

starwars %>%
  summarise(mean(species != "Human", na.rm=TRUE))
``` 


## Piping verbs

Since each verb requires a data frame and returns a data frame, we can combine verbs in a pipe.  
Here is an example.

```{r}
## average mass of humans 
starwars %>%
  filter(species == "Human") %>%
  summarise(av_m=mean(mass, na.rm=TRUE))

# without pipe
summarise(filter(starwars, species == "Human"), av_m=mean(mass, na.rm=TRUE))

temp1 <- filter(starwars, species == "Human")
summarise(temp1, av_m=mean(mass, na.rm=TRUE))

# with base R
temp2 <- starwars[starwars$species=="Human",]
av_m <- mean(temp2$mass, na.rm=TRUE)

av_m <- mean(starwars[starwars$species=="Human",]$mass, na.rm=TRUE)
av_m
```

Another example.

```{r}
# create the variable n_films for each name

# with base R
aa <- numeric(length(starwars$films))
for (i in 1:length(starwars$films)){
  aa[i] <- length(starwars$films[[i]])
}

starwars$n_films <- aa

# with dplyr 
c_l <- function(ll){
  aa <- numeric(length(ll))
  for (i in 1:length(ll)){
   aa[i] <- length(ll[[i]])
  }
  return(aa)
}

starwars %>%
  mutate(n_films = c_l(films)) %>%
  select(name, n_films) %>%
  filter( n_films >=5)

```

## `group_by`

The function `group_by` allows to make manipulations on groups.  
Let's make an example.

```{r}
starwars %>%
  group_by(eye_color) %>%
  summarise(
    n_per_group = n()
  ) %>%
  filter(n_per_group > 1)
``` 

## Keys for joins

`dplyr` also contains tools to join data sets. We have a glimpse at them here.  
Notice that joining data sets means forming a third data set with columns of a first and a second data set.  

A key is a column or a combination of columns.  
In order to join two data sets, two keys: a primary and a secondary key.  
The primary key must **uniquely identify the rows** of the primary data set; hence it can consist on many variables.  
The secondary key only needs to match the primary key.  

An example to illustrate this point uses the two following data sets `population` and `capital`.  

- If `population` was the primary data set, then the variable `country` alone could not be the primary key because it has repeated values.  

- On the other hand, if `capitals` was the primary data set, then the variable `country` could be that primary key.

```{r}
population <- tribble(
~country, ~year, ~population,
"UK", 2010, 62.7,
"UK", 2000, 58.9,
"FR", 2010, 65,
"FR", 2000, 60.9,
"RSA", 2010, 50.7,
"PT", 2010, 10.5
)

capitals <- tribble(
  ~country, ~capital, 
  "RSA", "Pretoria",
  "FR", "Paris",
  "UK", "London",
  "SP", "Madrid"
)

```

## Joins

Some joins are called mutate because they create a new data set: left_, right_, inner_ or full_ joins are of this type. Filter joins include semi_ and anti_join and they filter one data set.

inner_join():
Returning all rows from x where there are matching values in y, and all columns from x and y. Important: In the case of multiple matches between x and y, all combinations of the matches are returned.

left_join():
Returning all rows from x, and all columns from x and y. Rows in x with no match in y will have NA values in the new columns. Important: In the case of multiple matches between x and y, all combinations of the matches are returned.

right_join():
Returning all rows from y, and all columns from x and y. Rows in y with no match in x will have NA values in the new columns. Important: In the case of multiple matches between x and y, all combinations of the matches are returned.

full_join():
Returning all rows and all columns from both x and y. Important: In the case of no matching values, NA is returned for the one missing.

- `left_join` and `right_join`

The usage is, for instance,

```r
left_join(df1, df2, by=c("key1", "key2"))
# with pipe
df1 %>%
  left_join(
    df2,
    by=c("key1", "key2")
    )

```
This call will keep all the rows of `df1` and augment it with columns of `df2`. Notice that some `NA`s will be created and the rows on `df2` that do not match the primary key will not appear.  
`right_join` would do for `df2` what `left_join` join does for `df1`.

```{r}
capitals %>%
  left_join(
    population,
    by = "country"
    )

population %>%
  left_join(
    capitals,
    by= "country"
    )

```


- `inner_join` and `full_join` 

The call is similar to above.  
`inner` takes the rows that match in **both** datasets while `full` takes all the rows that match and those that do not match.

```{r}
capitals %>%
  inner_join(
    population,
    by = "country"
    )

capitals %>%
  full_join(
    population,
    by = "country"
    )

```


- `semi_join` and `anti_join`

These give a copy of the first data set filtered with the second data set. Hence, it's a way to filter data from the first dataset based on information in a second dataset.  
It can be used to quickly check which rows appear in both datasets

```{r}
capitals %>%
  semi_join(
    population,
    by = "country"
    )

capitals %>%
  anti_join(
    population,
    by = "country"
    )

```


## Operations on data sets

A few other functions allow to compare data sets. These include  the self-explanatory `union`, `intersect` and `setdiff`

`setdiff`:

```{r}
population2 <- add_row(population, country="GER", year=2015, population=10)

setdiff(population, population2)
setdiff(population2, population)
```

`intersect`:

```{r}
population2 <- add_row(population, country="GER", year=2015, population=10)

intersect(population, population2)
intersect(population2, population)
```

Other useful functions are `setequal` and `identical`.
Use `setequal(df1, df2)` which will be `TRUE` if the datasets have the same data even if the rows are not in the same order, opposed to `identical(df1, df2)`

`identical`:

```{r}
capitals <- tribble(
  ~country, ~capital, 
  "RSA", "Pretoria",
  "FR", "Paris",
  "UK", "London",
  "SP", "Madrid"
)

capitals2 <- tribble(
  ~country, ~capital, 
  "RSA", "Pretoria",
  "UK", "London",
  "FR", "Paris",
  "SP", "Madrid"
)

identical(capitals, capitals2)
```

`sequetal`

```{r}
capitals <- tribble(
  ~country, ~capital, 
  "RSA", "Pretoria",
  "FR", "Paris",
  "UK", "London",
  "SP", "Madrid"
)

capitals2 <- tribble(
  ~country, ~capital, 
  "RSA", "Pretoria",
  "UK", "London",
  "FR", "Paris",
  "SP", "Madrid"
)

setequal(capitals,capitals2)
```


