# `ggplot2` {#ggplot2}

ggplot2 is a system for creating graphics and therefore a system for data visualization. You provide the data to R, tell ggplot2 how to map variables and what graphical primitives to use, and it takes care of the details for you.

Data visualization in general is part of the skill set of a data scientist.  
It is made to, either
    - explore (confirm, analyse)
    - explain (inform, convince)

Depends who you communicate to (can be you).

To install the CRAN version:

```{r
install.packages("ggplot2")
```

## Comparison with base `plot`

Limitations of base plot:
    - plot does not redraw, i.e., the range will not adapt to new data 
    - plot is drawn as an image, i.e., it's not an object
    - manual legend
    - no unified framework for plotting, i.e., need to master other commands for other types of graphs

Example:

```{r, fig.show='hold', results='hide', tidy=FALSE}
library(ggplot2)
rm(mtcarts)
data("mtcars")
mtcars$cyl <- as.factor(mtcars$cyl) # treat 'cyl' as a factor

plot(mtcars$wt, mtcars$mpg, col = mtcars$cyl)
abline(lm(mpg ~ wt, data = mtcars), lty = 2)
lapply(mtcars$cyl, function(x) {
  abline(lm(mpg ~ wt, mtcars, subset = (cyl == x)), col = x)
  })
legend(x = 5, y = 33, legend = levels(mtcars$cyl),col = 1:3, pch = 1, bty = "n")
       
plot1 <- ggplot(mtcars, aes(x = wt, y = mpg, col = factor(cyl))) +
         geom_point() +
         geom_smooth(method=lm , se=FALSE, aes(col=cyl)) +
         geom_smooth(method=lm , se=FALSE, linetype=2, col="grey", aes(group=1))       
plot1 # print the object 'plot1'
```

## Grammar of graphics - Leland Wilkinson

### A graphic = layers of grammatical elements
The layers are the adjectives and the nouns
Seven grammatical elements, three are essential

### Meaningful plots are built around appropriate aesthetic mappings
The mappings are the rules to assemble the nouns and adjectives


## Understanding the grammar

Building an example:

```{r}

p1 <- ggplot(data=diamonds, mapping= aes(x=carat, y= price)) # data and mappings

#p2 <- p1 + geom_point()  # + the general form

p2 <- p1 + geom_line()  # + the general form

p3 <- ggplot(diamonds, aes(x=carat, y= price, col=clarity)) +
      geom_point() # adding a mapping / variable

p3b <- ggplot(diamonds, aes(x=carat, y= price)) +
      geom_point(col="red") # adding a attribute


p4 <- p1 + geom_point(aes(col=clarity)) # or changing some attributes
grid.arrange(p1, p2, p3, p4, ncol=2)
```

### Data and proper data format
The data being plotted
Includes: variables of interest
Tidy data helps make good `ggplot()`s

```{r, fig.show='hold', tidy=FALSE}
iris.tidy <- iris %>%
  gather(key, Value, -Species) %>%
  separate(key, c("Part", "Measure"), "\\.")

p1 <- ggplot(iris.tidy, aes(x = Species, y = Value, col = Part)) +
geom_jitter() +
facet_grid(. ~ Measure)

p2 <- ggplot(iris.tidy, aes(x = Measure, y = Value, col = Part)) +
geom_jitter() +
facet_grid(. ~ Species)


p3 <- ggplot(iris.tidy, aes(x = Species, y = Value, col = Part)) +
geom_point()

grid.arrange(p1, p2, p3, ncol=2)

```

### Aesthetics
The scales onto which we map our data
Includes: x-axis, y-axis, colour, fill, size, labels, line width, line type,...

### Geometries
The visual elements (shape) used for our data in the plot
Includes: point, line, histogram, bar, boxplot,...

### Other grammatical elements
These are
    - facets
    - statistics
    - coordinates
    - themes

### Juggling with `aes`
```{r}
data("mtcars")
p1 <- ggplot(mtcars, aes(x = wt, y = mpg)) +
      geom_point()

# color dependent on 'disp',
# 'disp' is continuous, hence the shades of same colour
p2 <- ggplot(mtcars, aes(x = wt, y = mpg, col = disp)) +
      geom_point()  + geom_smooth(se=FALSE)

# changing the colour depending on the variable 'clarity', 
# 'clarity' is a factor, hence the different colours
p3 <- ggplot(diamonds, aes(x = carat, y = price, col=clarity)) +
      geom_smooth()

# size dependent on 'disp'
p4 <- ggplot(mtcars, aes(x = wt, y = mpg, size = disp)) +
      geom_point()

grid.arrange(p1, p2, p3, p4, ncol=2)
```

### Juggling with `geom`

```{r,  fig.show='hold'}
# points and smoother layer
p1 <- ggplot(diamonds, aes(x = carat, y = price)) + 
      geom_point() + 
      geom_smooth(se=TRUE) # se = T by default

# only the smoother layer
p2 <- ggplot(diamonds, aes(x = carat, y = price)) + 
      geom_smooth()

# changing the parameters of the point geometry, much more transparent, here
p3 <- ggplot(diamonds, aes(x = carat, y = price)) + 
      geom_point(alpha=.04)

# only the smoother layer and different aesthetics,
# aes recognizes the groups and therefore separates for them
p4 <- ggplot(diamonds, aes(x = carat, y = price, col=clarity)) +
      geom_smooth() + geom_point(alpha=.04)

grid.arrange(p1, p2, p3,p4,  ncol=2)
```

## Aesthetics

### Understanding aesthetics
Usually considered as how something looks, as attributes: that's *notcorrect in `ggplot()` 
Aesthetics refers to what a variable is *mappedonto it
*Aesthetics is mapping*
We want to map as many variables as possible in a plot, that's visible aesthetics
Aesthetics/mappings are called in `aes()` while attributes are called in `geom_()`

For instance:
    - `aes(x = variable1, ...)`  means 'variable1' is mapped onto the x-axis
    - `aes(..., col = variable2)` means that 'variable2' is mapped onto a colour

Simply changing the colour of the dots, is NOT aesthetics
    - `ggplot(mtcars, aes(x = wt, y = mpg)) + geom_point(col="red")`
because there is no mapping
`aes()` includes: x-axis, y-axis, colour, fill, size, alpha, labels, line width, line type,...
Some aesthetics are only applicable to categorical variables: e.g., label and shape 

```{r, fig.show='hold'}
p1 <- ggplot(mtcars, aes(x = wt, y = mpg, col = cyl)) +
      geom_point(shape = 1, size = 4)
p2 <- ggplot(mtcars, aes(x=mpg, y=qsec,size=(hp/wt), 
                         shape=factor(am), col=factor(cyl))) + 
      geom_point()

grid.arrange(p1, p2,  ncol=2)
```

Attributes go to `geom_`

### Modifying aesthetics

Modifying aesthetics is modifying the mapping

`position` defaults to `identity`, `jitter` adds some random noise to the observations so that they do not overlap
```{r}

p1 <- ggplot(iris, aes(x=Sepal.Length, y=Sepal.Width, col=Species)) +
      geom_point()
p2 <- ggplot(iris, aes(x=Sepal.Length, y=Sepal.Width, col=Species)) +
      geom_point(position="jitter")

grid.arrange(p1, p2,  ncol=2)
```

`scale_..._...(variable1, args)` modifies the scale of the mapping of 'variable1' with several arguments such as `limits`, `breaks`, etc...
Sometimes, you need to assign a dummy to the aesthetics, e.g., when you want to plot only one variable

```{r}
mtcars$mydummy <- 1
 
ggplot(mtcars, aes(x = mpg, y=mydummy, col=factor(cyl))) + 
geom_jitter() + 
scale_y_continuous(limits=c(-0,2)) + 
labs(title="My nice graph", x="Miles per gallon", y="", col="Cylindres")
```

## Geometries

Geometries control how your plot is going to look like: currently, there are more than 37 gemoetries available
Common plots are: scatter, bar, line plots
Geometries required some aesthetics and can take other aesthetics as optional
Geometries can take their own aesthetics (useful to control mappings of each layer)
Geometries can take their own data, too!

### Scatter plots
We've seen a lot, already: recall `geom_point()`

### Bar plots
The simplest is a histogram
```{r}
p1 <- ggplot(iris, aes(x=Sepal.Length)) + geom_histogram()
p2 <- ggplot(iris, aes(x=Sepal.Length)) + 
      geom_histogram(binwidth = 1)
p3 <- ggplot(iris, aes(x=Sepal.Length, fill=Species)) +
      geom_histogram()
p4 <- ggplot(iris, aes(x=Sepal.Length, fill=Species)) +
      geom_histogram(position="dodge") # or "stack", or "fill"

grid.arrange(p1, p2, p3, p4,  ncol=2)
```

A more general bar plot can be obtained with `geom_bar()`:

```{r}
data("mtcars")
mtcars$cyl <- as.factor(mtcars$cyl) # changing 'cyl' to factor, which is actually is 
mtcars$am <- as.factor(mtcars$am)
p1 <- ggplot(mtcars, aes(x=cyl, fill=am)) +
      geom_bar(position="stack")
p2 <- ggplot(mtcars, aes(x=cyl, fill=am)) +
      geom_bar(position="dodge") # or "stack", or "fill"

posn_d <- position_dodge(width=.2) # like for jitter
p3 <- ggplot(mtcars, aes(x=cyl, fill=am)) +
      geom_bar(position=posn_d, alpha=.6)

p4 <- ggplot(mtcars, aes(mpg, fill=cyl)) +
      geom_histogram(binwidth = 1, position="identity", alpha=.4)


grid.arrange(p1, p2, p3, p4,  ncol=2)
```


```{r}
data <- data.frame(a=1:6, b=rep(1/6,6))

p1 <- ggplot(data, aes(x=a, y=b)) +
      geom_bar(position="stack", alpha = 0.7, stat = "identity") +
      labs(title = "Probability distribution of die roll", x = "Value on the die (x)", y = "P(x)") +
  scale_x_discrete(limits=1:6, labels=c("1", "2", "3", "4", "5", "6"))
      
p1
```

### Line plots

Particularly suited for line plots over time
```{r}
p1 <- ggplot(economics, aes(x = date, y = unemploy)) + 
geom_line()

data("beavers")
p2 <- ggplot(beaver1, aes(x = time, y = temp, col=factor(activ))) + 
geom_line(aes(group=1)) # group = 1 indicates you want a single line connecting all the points, you can set group=variable

grid.arrange(p1, p2, ncol=2)
```