# (PART) Use R for Data Science {-}

# Simple functions and operations on vectors 

This section gives simple calculations and examples for using vectors and their syntax in R language. The practical examples should help to achieve a general understanding.  


## Element by element evaluation

An important characteristic in relation to vectors is the use of functions. Each element of the vector is considered separately, evaluated and viewed in the sequence defined, also called **element by element**. 

The following example illustrates this evaluation and emphasizes the effect on the length of the vectors.
```{r}
# Creation of two numeric vector of same length (you can choose a random name)
shares1 <-  c(16, 23, 34, 32, 96, 22, 6)
shares2 <- c(11, 33, 2, 8, 36, 4, 69)
length(shares1)
length(shares1) == length(shares2)

total <- shares1 + shares1
# the sum of the vectors of the same length, with the element by element sums

length(total)
total

```

Another simple calculation with the function `^2`.


```{r}
total.p2 <- total^2
total.p2

(trop <- total - 2)
length(trop)
```

You also can run this with other functions respective exponents.


```{r}
total.p2 <- total^3
total.p2

(trop <- total - 3)
length(trop)
```

## Recycling rule


Now that we have examined functions in interaction with vectors of the same length, the question arises what happens if the selected vectors have a different length. What's important: the shortest vector has its elements **recycled** as much as it is necessary.

```{r}
l6 <- c(12, 34, 50, 8, 16, 15)
l4 <- c(10, 3, 13, 9)
tt <- l6 + 7
tt

ltotal <- l6+l4 # recycling takes place!
ltotal
```


## Functions for the whole vector

Functions do not necessarily have to refer to vectors of different or equal length. With R various operations on vectors can be made.
Here are a few simple examples to illustrate this:

```{r}
coins <- c(36, 73, 16, 98, 22, 23, 87, 19, 12)
coins
max(coins)
length(coins)
sum(coins)
var(coins)

coins <- c(36, 73, 16, 98, 22, 23, 87, 19, 12)
mean(coins)

# for help on a command, simply type ? in front 
# of unknown command to search for explanation
?var
```