# Import data to R {#import}



Analyzing data with R is an important part of data science. Therefore, it is vital to understand the procedure of how to import data into R. Generally speaking, the process of importing data into R is very simple. Data out there exists in many formats and there are number of ways in importing data into R. Speakin the language of R, we will need to `read` the data and put or convert it into an object, wich often is a data frame. 

Base R has has base functions to read some types of files. For special types of files, special packages are needed.

**Important note:** these base functions are introduced here for documentation purpose only. 

## Reading rectangular data

Rectangular data is intended here as a data set with values of variables in columns and each row representing a case. Arguably, this is the simplest and probably most common way to store data sets.  
The ad hoc function in R is `read.table`. Other similar functions exist, e.g., `read.csv`, but they are essentially a wrapper of `read.table`.   
This function reads rectangular data and assigns it to an object, in almost all cases a data frame. The arguments of the function are numerous are, depending on the case, potentially very useful (check `?read.table`).  We mention here a few of them in the typical call:

```{r
my.data <- read.table(`file`)
``` 

`file` is the name of the file to be read. It requires a valid path. It can also be an `url`, with some adaptations.  
`header` is a logical indicating whether the data's first line gives the names of the variables.  
`sep` indicates the character that separates the values between columns.

```{r}
# assuming we have the file AH001.txt in the same folder
df <- read.table(file="AH001.txt",
				header=TRUE,
				sep=",")
head(df)
```


## Read other data types

R can import virtually all types of data. For special types, a ad hoc package will be necessary. `foreign` is one of them and it allows to import data from other applications. Its functions are of the form

```r
read.XXX
```
 where the `XXX` is an extension specific to the external software. Examples include "read.dta" for Stata data files, `read.spss` for SPSS files or `read.dbf` for dBASE files.   

A very important data source type is Excel. Functions for this type of data are covered in Chapter \@ref(readr) from the part on `tidyverse`.

Very often, the data source is in an Excel format, and needs to be imported into R prior to using it. Therefore, we can make use the function read.xls from the gdata package (Install the package: _Tools> Install packages_). It reads the data from an Excel spreadsheet and returns a data frame. 

The following example is an example of how to load an Excel spreadsheet named "datascienceproject.xls" into R (This method requires Perl runtime to be present in the system):

```{r
library(gdata)                   # loading the above mentioned gdata package 
help(read.xls)                   # documentation (help regarding read.xls) 
mydata = read.xls("datascienceproject.xls")  # read from the excel spreadsheet
```

An alternative solution would be to make use of the function loadWorkbook from the XLConnect package to read the entire workbook, and then load the worksheets with readWorksheet (The XLConnect package requires the software Java to be pre-installed).

```{r
library(XLConnect)               # loading the above mentioned XLConnect package 
dsp1 = loadWorkbook("datascienceproject.xls") 
dsp2 = readWorksheet(dsp1, sheet="Sheet1")                # Example of reading sheet 1 of the entire worksheet.
```

SPSS File
For data files in the format of SPSS, it can be opened with the function read.spss, which is also part of the foreign package. There read.spss function also includes a "to.data.frame" option, in order to choose whether a data frame is to be returned. By default, it returns a list of components instead.

```{r
library(foreign)                 # loading the above mentioned foreign package 
help(read.spss)                  # documentation (help regarding read.spss) 
mydata = read.spss("datascienceproject", to.data.frame=TRUE)
```

CSV File

The sample data can also be in comma separated values (CSV) format. Each cell inside such data file is separated by a special character, which usually is a comma, although other characters can be used as well.

The first row of the data file should contain the column names instead of the actual data. Here is a sample of the expected format.

```{r
Col1,Col2,Col3 
100,a1,b1 
200,a2,b2 
300,a3,b3
```

After we copy and paste the data above in a file named "datascienceproject.csv" with a text editor, we can read the data with the function read.csv.

```{r
datascienceproject = read.csv("datascienceproject.csv")  # read the csv file 
mydata 
  Col1 Col2 Col3 
1  100   a1   b1 
2  200   a2   b2 
3  300   a3   b3

help(read.csv)            # If help is required for the read.csv function
```

Working Directory

Finally, the code samples above assume the data files are located in the R working directory, which can be found with the function getwd.

```{r
getwd()               # get current working directory
```

You can select a different working directory with the function setwd(), and thus avoid entering the full path of the data files.

```{r
setwd("<new path>")   # set working directory
```

Note that the forward slash should be used as the path separator even on Windows platform.

```{r
setwd("C:/MyDoc")
```



## Scanning a file

The base function `scan` can sometimes be useful, though it is not used to import data. `scan()` imports to a vector (or a list), which can then be used.   

- `what` gives the type of vector to be imported.  
- `skip` is the number of lines to be skipped in the file.  
- `nlines` determines the number of lines to be read and imported.

```{r}
df <- scan("AH001.txt", what=character())
#df
df <- scan("AH001.txt", what=character(), skip=1, sep=",")
#df
given.line <- scan("AH001.txt", what=character(), nlines=1, sep=",")
given.line
```





