# Subsetting {#subset}


There are three subsetting operators: `$`, `[]` and `[[]]`.  
Some functions also allow to create subsets. 
We can combine subsetting and assignment to change some parts of an object.  
Sub-setting is better used in complement with `str()`.
```{r}
va <- c(13.1, -15.2, 0.3, 2.4, 10.5, -3.6, 9.7) # create the vector va
str(va)
str(mtcars)
```

One can see from this call, that `va` is a simple vector with `r length(va)` elements. Subsetting means choosing among these.

## `[]`
Applies to vectors, matrices, lists and data frames.  
Can be used with:

    - positive or negative values,
    - many values in a vector,
    - logical, `NA`,
    - character vectors when names exist.

### On a vector
Here are a few examples of that object used on a vector.

```{r} 
va <- c(13.1, -15.2, 0.3, 2.4, 10.5, -3.6, 9.7)
va[1] # element 1
va[c(3:5)] # elements 3 to 5
va[c(2,7)]
va[-1] # all elements minus the element 1
va[c(-2,-7)]
va[c(-2,-7)]
va[c(TRUE,TRUE,TRUE,FALSE,FALSE,TRUE,FALSE)] 
va[c(TRUE,FALSE)] # notice the recycling at play here
va[NA] 
va[] # nothing selected gives the full vector
names(va)<-letters[1:length(va)] # give names to va
# notice that we subset the vector letters (given by R) and that we don't
# specify the length but give a general value
# now we can subset using names
va
va[c("a","e","b")]

```
### On a list
```{r}
mylist<- list(numbers=c(1:20), 
              ournames=c("Jim","Jules"), 
              results= c(T,F,F,T), 
              school=factor(c("Primary", "Secondary", "Tertiary"), ordered=TRUE))
str(mylist)

mylist[1] # first element of the list
class(mylist[1])
mylist[[3]][3]
```
Notice that the result of this subsetting is a `r class(mylist[1])`!

### On a matrix

```{r}
set.seed(42)
my.mat <- matrix(floor(runif(30)*10), nrow=5)
my.mat
str(my.mat)
length(my.mat)
```
The structure shows that there are r length(dim(my.mat)) dimensions. For subsetting, we must give r length(dim(my.mat)) dimensions! Same rules as for the vectors apply.

```{r}
# 2nd row, 3rd column
my.mat[2,3] 
# all but first row, all columns
my.mat[-1,]
colnames(my.mat) <- letters[1:ncol(my.mat)] # give names to the columns
rownames(my.mat) <- LETTERS[1:nrow(my.mat)] # give names to the rows
my.mat
my.mat["C",c("a","c","e")]
```

## `[[]]`
This object is used mainly for lists.

```{r}
mylist<- list(numbers=c(1:20), 
              ournames=c("Jim","Jules"), 
              results= c(T,F,F,T), 
              school=factor(c("Primary", "Secondary", "Tertiary"), ordered=TRUE))
str(mylist)
mylist[[1]]
```


## `$`
This object is usually for data frames, where it gives the variable.  
It allows partial matching (e.g., `mtcars$gear` is the same as `mtcars$gea`)

```{r}
data(mtcars)
mtcars
mtcars[1:5, 4]
aa <- mtcars[1:5, c("cyl","hp")]
class(aa)
aa

names(mtcars)
mtcars$hp[1:5]

mtcars[["mpg"]] # just exactly the same, but R users prefer the $
mtcars["mpg"] # NOT the same at all! [] preserves the class
class(mtcars["mpg"])
```

## Combining subsetting
Notice that we can often subset further until having the desired subset. Here are a few examples.
```{r}
mtcars$mpg[1:4]
mylist[["ournames"]][1]
mylist$ournames[1]

```



## Subsetting  with one condition
We can use conditions for subsetting
```{r }
mtcars[mtcars$mpg<=20, ]
mtcars[mtcars$cyl==8,]
mtcars[mtcars$cyl==8 & mtcars$carb==4,]
mtcars[mtcars$cyl==8, c("cyl", "mpg", "wt")]

```

## Subsetting and assignment
Subsetting can be used to change a part of an object through assignment.  Assign `NULL` to delete subset
```{r}
my.mat
my.mat[,1]<-1:nrow(my.mat)
my.mat
mtcars$mpg[1]<-1234
names(mtcars)
mtcars$drat<-NULL # delete variable drat in data frame mtcars
# does not delete in matrix
names(mtcars)

```



## Using `which()`
`which()` gives the integers that correspond to the boolean (logical) `TRUE`.  
This can help subsetting
```{r}
vb<-1500:1530
vb
which(vb%%5==0) # which is divisible by 5 (modulo is 0) ?
# notice that this is asking each element of vb, 
# which() reports the positions for which the answer is TRUE 
vb[which(vb%%5==0)] 
```

## More advanced stuff

### Difference between simplifying and preserving
We say 'preserve' to say the same structure is maintained when subsetting (e.g., a subset of a data frame remains a data frame).  
Simplifying does not keep the structure but gives the simplest output possible.  
`drop` argument allows to preserve (`drop=FALSE`) or not (`drop= TRUE`).  
`[]` usually preserves, `[[]]` usually simplifies.  
To better understand, check the classes:
```{r}
all.equal(class(mylist[1]), class(mylist[[1]]))
class(mylist[1])
class(mylist[[1]])
all.equal(class(mylist), class(mylist[1]))
all.equal(class(mylist), class(mylist[[1]]))
```

We see that `[]` has preserved the class of `mylist` (r class(mylist)), while `[[]]` has not! 
Another striking example is the following.
```{r}
ma2 <- my.mat[1:2,1:3]
class(ma2)
ma3 <- my.mat[1,1:3]
class(ma3)
```
`ma3` is NOT a matrix anymore!!! This is because one of its dimensions is 1.
Losing track of the class when subsetting can generate lots of problems in the middle of a large code. And it is a common source of error!
To be sure of keeping the class, we can use `drop=FALSE` (the class will not be dropped to, usually, a vector).
```{r}
ma4 <- my.mat[1,1:3, drop=FALSE]
ma4
class(ma4)
```

