---
output:
  pdf_document: default
  html_document: default
---
# Conditions {#conditions}

The general intention and reason for conditions is to control the flow of the calculation and output when executed by R.  
With R language you can choose between a variety of conditions, but most common are `if` and `else`.

## `if` statement
The most simple conditions uses a `if` statement.  
The form is then:

```r 
if (the definded condition) {
# code to be executed if the condition is TRUE
}
```

It is important to point out here that R executes the code until a condition is identified, otherwise R runs through the next line in the chunk.

Here is a simple example uses this condition:
```{r}
losses <- c(13, 5, -60, 0, -7, 24, 8)
losses
sum(losses)

if (sum(losses) < 0) {
  print("Biggest looser in town")
}

# change the -60 in 'losses' to -6

losses[losses==-60] <- -6
losses

if (sum(losses) < 0) {
  print("Biggest looser in town")
} # remember that there is no print or output, because the defined condition was not met
```

## `else` statement

What can be the case if the condition was not met? It can be no output or... something else!
It is up to you to decide what you want to do in this situation if the first condition is not met!

```r
if (condition) { 
# code to be executed if the condition is TRUE
} else {
# code to be executed if the condition FALSE
}
```

To illustrate:

```{r}
#we take the same example as above
losses<- c(13, 5, -60, 0, -7, 24, 8)
losses[4] <- 50 # change the fourth element of losses
sum(losses)

if (sum(losses) < 0) {
  print("Biggest looser in town")
} else {
  print("Lucky Cowboy")
} # notice that now there is an output!
```

## `else if` statement

`else if` allows to work with another condition additional to the current flow of code above

```r
if (condition_1) { 
# code to be executed if the condition_1 is TRUE
} else if (condition_2) {
# code to be executed if the condition_2 is TRUE
} else {
# code to be executed if NEITHER condition_1 NOR condition_2 is TRUE
}
```

There is no limitation in `else if` statements. Many multiple conditions are possible  
Here is an example with only one `else if`
```{r}
losses <- c(13, 5, -60, 0, -7, 24, 8)
losses[3] <- 25 # change the third element of 'losses'
sum(losses)

losses <- c(13, 5, -60, 0, -7, 24, 8)
losses[6] <- 41 # another one
sum(losses)

if (sum(losses) < 0) {
  print("Biggest looser in town")
} else if (sum(losses)==0) {
  print("No looser, but no champion at all")
} else {
  print("Lucky Cowboy")
}
```

Pay attention the potential problem of not putting an  `else` statement at the end of the conditions system.

## `ifelse` statement

When using **shorter** conditions, we can use `ifelse` statements.  
Here is how the code looks like:

```r
ifelse(condition, value if condition met, value if value not met)
```
Now back to our real example:

```{r}
losses <- c(13, 5, -60, 0, -7, 24, 8)
sign <- ifelse(sum(losses)>=0,"+","-")
sign

?sign #sign returns a vector with the signs of the corresponding elements of x (the sign of a real number is 1, 0, or -1 if the number is positive, zero, or negative, respectively)
```

## Logical operators: `&`, `|` and `!`
Logical operators are used to combine, mix or negate several conditions:
- `&` means AND
- `|` means OR
- `!` means NOT

In Boolean algebra, there are two transformation rules named after the British mathematician Augustus De Morgan. These rules state that:

1. The complement of the union of two sets is the intersection of their complements.
2. The complement of the intersection of two sets is the union of their complements.

De Morgan's laws may help here.

```{r}
(27%%2==0 & 27%%3==0) # equivalent to (TRUE and TRUE), hence TRUE
TRUE & FALSE
!TRUE 
!(TRUE & TRUE)
!(TRUE & !TRUE)
((27%%2==0 & 27%%2==0)) # equivalent to (TRUE and FALSE), hence FALSE
((27%%2==0 | 27%%2==0)) # equivalent to (TRUE or FALSE), hence TRUE
```


## Writing and interpreting a condition

It must be considered that R searches either a TRUE or a FALSE in a condition.  
When R encounters a TRUE, it executes the desired command, otherwise it does not.  
There are many ways to get one of these two logics, since each of these ways will work as a condition.

Here are examples of less common ways to write a condition:
```{r}
losses<- c(13, 5, -60, 0, -7, 24, 8)

if (is.numeric(losses)) {
  print("Wow a wild numeric vector...")
}
# here, is.numeric(losses) evaluates to TRUE, hence, the code is executed!
# the beginner's way would be to replace the condition by
# if (class(losses)=="numeric")

if (!is.factor(losses)) {
  print("No factor no portuguese friday...")
}
# the amateur would write if (class(losses)!="factor")
```

