# Functions {#functions}

Everything that happens in R is a function call. If we want R to do something, we must use a function, may it be as simple as a parentheses `(`. 
If the function is not built-in or coming in a package, one must write it.
A function is itself an object. When applied to another object, the function "makes something" to that object and returns an object.

## Structure of a simple function

Here is a simple example of a function that returns the square of the object provided.

```{r}
pow_two <- function(a){
  a^2
}
pow_two(12)

```



### Name of the function
In our example above, the name was `pow_two`. To call, i.e., to use the function later in the code, type `pow_two()`.

### Arguments

The arguments are given in the parentheses right after `function`.  
Here, we use an argument `a`, which can be any object, e.g., a vector
```{r}
pow_two(1:10)
```
When using this function, we can see that one argument only must be passed.  
A value for the argument is then given in the parentheses when calling the function, e.g., `pow_two(12)`.  
A function can have several arguments

### Commands
The commands will be executed, if possible.  
One command source of error is when the class of the argument provided is NOT compatible with the commands in the function

```{r , error=TRUE}
pow_two("hello") # this can't work
```


### Return of a function

The return of a function is given by the last expression executed inside the function.

```{r }
my_f <- function(a){
  a^2
  a^3 # THIS is the last expression executed
}
my_f(5)
```

We can use `return()` in the function in order to chose what it  returns.

```{r }
my_f2 <- function(a){
  return(a^2)
  a^3
}
my_f2(5)
```

A function can return only one object; this is not really a  problem as we  can always stack everything in a `list()` and have the function return the list.

```{r }
my_f3 <- function(a){
  list(ptwo=a^2, pthree=a^3, pfour=a^4) 
}
my_f3(5:20)
```

If we want to use the return of our function, we must assign it to an object.

```{r }
x <- my_f3(1:5) # evaluates the function above on the vector 1:5
x$pthree # shows the element pthree of the saved list x
```

An example where we better use `return` is the following

```{r}
pow_three <- function(a){
  if (!is.numeric(a)) {
    return(print("You must give a numeric vector!"))
  } else {
    return(a^3)
  }
  
}
pow_three(1:4)
```



## Multiple arguments and their identification

```{r}
pow <- function(a, p){
  a^p
}
pow(12, 0.5)
```


As mentioned above, a function can have several arguments.

```{r , error= TRUE}
my_f4 <- function(abc, xyz){
  c(abc^2, xyz^3)
}
my_f4(5) # this won't work because xyz is missing
my_f4(5, 10)
```

R has at least two ways of identifying arguments, by position and by name.  
If no name is specified, R uses the arguments according to their positions in the definition of the function.  
In the example above, R understands that `abc=5` and `xyz=10` because of the way they are given in the call `my_f4(5, 10)`.  
Identification by names would be
```{r , error= TRUE}
my_f4(5, 10)
my_f4(abc=5, xyz=10) # same as before
my_f4(xyz=5, abc=10) 
# different from before because the names are more important than the position
my_f4(xyz=5, 10) # a mix between name and position 
```

### Default values

A function can be defined with default values for their arguments.  
R will use the values provided in the call; if one is missing, R will use the default value.

```{r , error= TRUE}

my_f4 <- function(abc, xyz=2){
  c(abc^2, xyz^3)
}
my_f4(5) 

my_f5 <- function(a, tada){
  if (tada==TRUE){
    plot(a)
  } else {
    sum(a)
  }
}


my_f5(1:20, tada=FALSE)

my_f6 <- function(a, tada=TRUE){
  if (tada==TRUE){
    plot(a)
  } else {
    sum(a)
  }
}

my_f6(1:30, tada=FALSE)
```












