---
output:
  pdf_document: default
  html_document: default
---
# Loops (avoid them) {#loops}



## Introduction to loops with two examples

Whenever you learn about R (e.g in a book or tutorial), the advice given on loops is always:

1. Learn about and try to get a clear understanding of loops, due to the fact that they provide you with an understanding of the data that you’re manipulating.

2. After you have managed to get a clear understanding of loops, get rid of them or try to avoid them whilst using R for data science.

According to the R base manual, among the control flow commands, the loop constructs are:
- for
- while
- repeat
- break (additional clause)
- next (additional clause)

Loops allow to iterate a set of functions / lines of code over a predefined list of elements.  
They are common in many languages but are not highly regarded in R because:

  - they do not represent an efficient way of producing a result,
  - functionals, functions that return a vector such as the `apply` family, are preferred.
    
This section serves as a short introduction to the functionals because it helps understand the structure of the problem (Given by two examples):  

## A silly example of loops (Bad R Coding)

Suppose our gains at a game over the week are, in euros:
```{r}
gains <- c(5, 2,-10,0,-2,15,1)
gains
```

Now, suppose we want to transform each gain into dollars, i.e., we must multiply each gain by 1.30. An intuitive way would be to take each gain _separately_ and multiply it 1.12. Let's do it by hand:

```{r}
gains_d <-numeric(length(gains)) # create a numeric vector of same length as gains
gains_d[1] <- gains[1]*1.12 
gains_d[2] <- gains[2]*1.12
gains_d[3] <- gains[3]*1.12
gains_d[4] <- gains[4]*1.12
gains_d[5] <- gains[5]*1.12
gains_d[6] <- gains[6]*1.12
gains_d[7] <- gains[7]*1.12
gains_d
```

We recognize that the structure of each line of code: the first line is for the element 1 of the vectors, the second line for the second element of the vectors, etc.  
Doing a loop builds on that observation. But it asks the program to make the change between each line automatically.  
In the silly example, we want the program to evaluate a line with the value 1 and then, when it has finished, do the same line but with a 2 instead of a 1. And we want it to do it **for** the elements 1, 2, 3, 4, ..., `r length(gains)`.  
This is what a loop does!


```{r}
# create an empty numeric vector of same length as gains
gains_d2 <-numeric(length(gains)) 

# now the loop
# first for i=1, then i=2, then i=3, etc... until i=length(gains)
for (i in 1:length(gains)) {  
  gains_d2[i] <- gains[i]*1.12 
}
gains_d2
```


## A good example of a loop

Suppose we want to know, every day, what is our cumulative gain until and including that day. This implies that we must to create another `cumulative_gains` with the same length as `gains`. 

Then, we could run the following loop.
```{r}
# create an empty vector
cumulative_gains <-numeric(length(gains)) 

cumulative_gains[1] <- gains[1]  
 
for (k in 2:length(gains)){
  cumulative_gains[k]<-cumulative_gains[k-1] + gains[k]
}
gains
cumulative_gains
```

