---
output:
  pdf_document: default
  html_document: default
---
# Data stuctures {#datastructures}

This chapter lists the most common objects to store data.


## Atomic vectors
The basic data structure in R is the vector. Vectors have three characteristics:

  - a type, `typeof()`,
  - a length, `length()`,
  - some attributes, `attributes()`. 

### Types of data 

There are four common types of atomic vectors:

  - logical,
  - integer,
  - double (often called numeric),
  - character.

Note that if vector elements not all of the same type, R makes coercion.  In that case, the order becomes:
"logical < numeric < character".

Here are a few illustrations.


```{r}
# logical vector
log_vector <- c(TRUE, FALSE, FALSE)
log_vector

# numeric vector
num_vector <- c(12, 10, 3)
typeof(num_vector)

# integer vector
int_vector <- c(12L, 10L, 3L)
typeof(int_vector)


# character vector
chr_vector <- c("a", "b", "c")
typeof(chr_vector)

# mixed types vector
mix_vector <- c(2, "a")
is.numeric(mix_vector)
typeof(mix_vector)
```

Numeric vectors can be created with the shortcut `:`, which is used to imply all the integers from the number on the left of `:` to the number on its right. Even if it is not good practice, these do not require the `c()` call. 

```{r vectors}
vector_A <- 1:5
vector_A
length(vector_A)
```

### Factor vector

This is a special type of vector. It has a limited number of values, called levels. These levels can be unordered (e.g., gender is either "Female" or "Male") or ordered (e.g. school level is "Primary", "Secondary", "Tertiary")
```{r factor}
gender <- factor(c("Male", "Male", "Female", "Male", "Female", "Female", "Male"))
gender
levels(gender)
summary(gender)
school <- factor(c("Primary", "Secondary", "Tertiary"),
					ordered=TRUE)
school

school2 <- factor(c("Primary", "Secondary","Secondary", "Tertiary"),
					labels=c( "Secondary", "Tertiary","Primary"),
					ordered=TRUE)
school2

```

## Matrices and arrays

R is not often used for matrices calculations: it is too slow for that and there are better programs for it out there (e.g. Matlab).

```{r}
# populate a matrix with the elements in of the vector, and give the dimensions
M <- matrix(c(4, 1, 0, 3, 6, 8), nrow=3, ncol=2) 
M
```

If we think of a matrix as a 2 dimensions vector, then arrays are $n$ dimensions vectors. Is it important? It might be in some very specific cases.

```{r array example}
mya <- array(data=1:18, dim=c(2,3,3))
mya
```

## Lists
These are the one-size fit all structure... A list is an object composed of any other object, even... another list! Very useful data structure!

```{r list example}
school<-factor(c("Primary", "Secondary", "Tertiary"), ordered=TRUE)
mylist <- list(numbers=c(1:60),
				somenames=c("Jim","Jules"),
				results= c(T,F,F,T),
				school=school)
mylist

# change the names of the elements of the list
names(mylist) <- c("N", "O","R","S")
```

## Data frames

Data frames are the second most important data structure in R. We can think of it as a better version of a data set in Excel. It stacks together observations over many variables, each of these variables being a vector.

```{r dataframe}
# mtcars is a built-in data set
data(mtcars)
class(mtcars)
mtcars
head(mtcars)
str(mtcars)

# names of the variables in the data frame
names(mtcars) 
length(mtcars)
nrow(mtcars)
```
### Data frame creation

A data frame can be created manually by providing the function `data.frame` with the corresponding vectors. These vectors should be of same length, otherwise R recycles values. The data frame requires a name for each vector.

```{r}
df <- data.frame(let=LETTERS[1:7],
                 num1=10:16, 
                 num2=floor(rnorm(7,100,10)))
df
names(df)
```

```{r}
students <- data.frame(name= c("tr", "ga", "mi", "st", "eb", "ha", "fo", "fi"),
                       age=c(23, 23, 24, 33, 28, 24, 33, 41),
                       like= c(T, T, T, T, F, F, T, T)) 
students
```

## A word about attributes

Attributes are important characteristics of vectors (and other data structures).   
They can be seen as some meta data that defines the nature of the object. This is to say that a given attribute can determine how a function applies to an object. Here are three important attributes:

  - the names, `names()`,
  - the dimensions, `dim()`,
  - the class, `class`.

To illustrate the use of attributes, recall the case of a factor vector. The fact that the vector on which it builds is given the attribute of a class (factor) turns that vector into a special one.

```{r}
vec <- c(1.7, 1.3, 4, 3.3, 3.3, 2, 2.3, 2.3)
vec
table(vec)

vec.f <- factor(vec,
                levels=c(1, 1.3, 1.7, 2, 2.3, 2.7, 3, 3.3, 3.7, 4, 5),
                ordered=TRUE)
vec.f
table(vec.f)
```


## Environments

An environment is similar to a bag/list of names. It can be seen as the space were these names "live".   
Environments can include environments. In that case, the enclosing environment is called the parent environment.  
The most common environment is the `globalenv()` also called the `workspace`. 


