---
output:
  pdf_document: default
  html_document: default
---
# Simple plots {#splots}

The following chapter is an introduction for producing simple graphs with the R Programming Language. The package `ggplot2` described in Chapter \@ref(ggplot2) allows to make much richer and elegant plots. For every plot explained, an example will be given. In this chapter, we will only focus on line plots, bar graphs, and histograms (other examples such as boxplots, pie charts or dotcharts are only mentioned and not explained with own examples).

## Line plots

The simplest plot is a scatter plot of one variable plotted against an index (1, 2, 3, ...) on the horizontal axis.  
In order to (scatter) plot one vector against the other, it is compulsory that they are both numeric and have the same length.  
The following plots illustrate the creation of a line plot in incremental steps. In the steps, the following functions and arguments are introduced:  

- `plot` is the main function for the plot.


```{r, echo=FALSE, include=FALSE, eval=FALSE}

d.plot <- data.frame(a = c("type", "col"),
                    d = c("p, l, c, o, s, h, n", "character"),
                    w = c("How the vector should be rendered in the plot in the plot",
                    "Determines the color; see this [Rcolor.pdf online](http://www.stat.columbia.edu/~tzheng/files/Rcolor.pdf) for details.") )
names(d.plot) <- c("Argument", "Value", "Description")
knitr::kable(d.plot, caption = '`plot` arguments.')                    
```

- `type` argument type gives how the vector should be rendered in the plot; the `type` of the plot is any of Table \@ref(tab:table-p-types). 


```{r table-p-types, echo=FALSE, align="c"}
p.types <- data.frame(Symbol = c("p", "l", "b", "c", "o", "s", "h", "n"),
                      Description=c("Points", "Lines", "Both", "The lines part alone of b", 
                      "Both overplotted", "Steps", "Histogram like (vertical lines)", "No plotting"))

knitr::kable(p.types, caption = 'Plot types.')
```
  
Example 1:

```{r simple-plots-1, out.width="25%", fig.align="center", fig.height=7, fig.cap="Creating a simple line plot.", fig.show="hold"}
sales <- c(20, 18, 24, 36, 30)
date <- c(15, 16, 17, 18, 19)
# left
plot(sales)
# middle
plot(date, sales)
# right
plot(date, sales, type="l")
```

Example 2:

```{r}
tarsus <- readxl::read_excel("tarsus.xlsx")
# left
plot(tarsus$Weight)
# middle
plot(tarsus$TarsusLength, tarsus$Weight)
# right
plot(tarsus$TarsusLength, tarsus$Weight, type="l")
```

We now add further customization with new functions and arguments.  

- `col` determines the color; see this [Rcolor.pdf online](http://www.stat.columbia.edu/~tzheng/files/Rcolor.pdf) for details.
- The `lty` argument refers to the type of line, to be chosen from the values shown in Figure \@ref(fig:line-types),
- `lines` adds the plot of a vector to a previously opened plot.
- `axis` is a function that changes the axis given in its first argument with 1, 2, 3, 4 referring to the bottom, left, top, and right axis, respectively.
- `at` simply states for what values of the axis the labels should correspond.
- `las` gives the orientation of the labels (e.g., `1=horizontal`).
- `xlab` and `ylab` are the x and y axes labels, respectively.
- `xlim` and `ylim` set a numerical limit for the x and y axes labels, respectively, notice that a vector of length 2 is necessary for each.


```{r line-types, echo=FALSE, out.width="50%", fig.height=3, fig.align="center", fig.cap="Line types."}
# adapted form http://www.sthda.com/english/wiki/line-types-in-r-lty
generateRLineTypes<-function(){
  oldPar<-par()
  par(font=2, mar=c(0,0,0,0))
  plot(1, pch="", ylim=c(0,3), xlim=c(0,0.5),  axes=FALSE,xlab="", ylab="")
  for(i in 0:6) lines(c(0.2,0.5), c(i/2,i/2), lty=i, lwd=3)
  text(rep(0.1,6), (0:6)/2, labels=c("0.'blank'", "1.'solid'", "2.'dashed'", "3.'dotted'",
                                 "4.'dotdash'", "5.'longdash'", "6.'twodash'"))
  par(mar=oldPar$mar,font=oldPar$font )
}
generateRLineTypes()
```



```{r simple-plots-2, out.width="25%", fig.align="center", fig.height=7, fig.cap="Custom a simple line plot.", fig.show="hold"}
sales2 <- c(12, 19, 22, 28, 32) 
# left
plot(date, sales, type="l", col="blue")
lines(date, sales2, type="b", col="red", lty=3)
# middle
s.range <- c(min(sales,sales2),max(sales,sales2))
plot(date, sales, type="l", col="blue", ylim=s.range)
lines(date, sales2, type="b", col="red", lty=3)
# right
plot(date, sales, type="l", col="blue",
     ylim=s.range,
     axes= FALSE, 
     xlab = "Day",
     ylab= "Sales")
lines(date, sales2, type="b", col="red", lty=3)
axis(1, at=date, lab=c("Mon","Tue","Wed","Thu","Fri"))
axis(2, las=1, at=seq(min(s.range),max(s.range),5)) 
```



We now add further elements such as a legend or a title.

- `legend` gives and places a legend; its first arguments are the location on the x and y axis respectively (this can be replaced by expressions such as `topleft`); the third is the legend itself, completed with colors and other formatting; notice that since the legend specifies two series, the formatting must also be given for each of them in vectors of length 2.
- `pch` determines the shape of the point in the plot, to be chosen from Figure \@ref(fig:point-types).
- `cex` gives the expanding factor of the text.
- `bty` specifies whether of not the legend should be in a box (default, "o") or not "n".
- `title` gives the title to the plot.


```{r point-types, echo=FALSE, out.width="50%", fig.height=3, fig.align="center", fig.cap="Point types."}
# adapted from http://www.sthda.com/english/wiki/r-plot-pch-symbols-the-different-point-shapes-available-in-r
generateRPointShapes<-function(){
  oldPar<-par()
  par(font=2, mar=c(0.5,0,0,0))
  y=rev(c(rep(1,6),rep(2,5), rep(3,5), rep(4,5), rep(5,5)))
  x=c(rep(1:5,5),6)/2
  plot(x, y, pch = 0:25, cex=1.5, ylim=c(1,5.5), xlim=c(1,3), 
       axes=FALSE, xlab="", ylab="", bg="blue")
  text(x, y, labels=0:25, pos=3)
  par(mar=oldPar$mar,font=oldPar$font )
}
generateRPointShapes()
```



```{r simple-plots-3, out.width="50%", fig.align="center", fig.height=7, fig.cap="Further customization of a simple line plot.", fig.show="hold"}
plot(date, sales, type="l", col="blue", ylim=s.range,
      axes= FALSE, xlab = "Day", ylab= "Sales")
lines(date, sales2, type="b", col="red", lty=3)
axis(1, at=date, lab=c("Mon","Tue","Wed","Thu","Fri"))
axis(2, las=1, at=seq(min(s.range),max(s.range),5)) 
legend(date[1], s.range[2], c("Tom","Mark"), cex=1, 
   col=c("blue","red"), pch=c(NA,1), lty=c(1,3), bty="n")
title(main="Week Sales", col.main="black", font.main=2)
```


One can help but notice how cumbersome the coding of the graph is, in particular when one needs to retype the whole code for the same graph.



## Bar graphs and histograms


This subsection illustrates the creation of these graphs with a few commented examples.

- `barplot` takes a vector or a matrix as first argument. It is therefore important to double check how this latter is created.
- `density` applies to the colors and a single value would be recycled.
- `horiz` defaults to `FALSE` and gives the direction of the plot.
- `border` applies to the borders of the bars.
- `beside` forces side-by-side bars instead of stacking bars.

Advanced example of a bar plot with additional functions:

```{r bar-plots, out.width="25%", fig.align="center", fig.height=7, fig.cap="Simple bar plots.", fig.show="hold"}
barplot(sales,
        main="Week Sales - Tom",
        names.arg=c("Mon","Tue","Wed","Thu","Fri"),
        border="lightblue",
        col = "blue")
barplot(sales,
        main="Week Sales - Mark",
        names.arg=c("Mon","Tue","Wed","Thu","Fri"),
        border="red",
        col = "red",
        horiz = TRUE)

barplot(matrix(c(sales, sales2), ncol=5, byrow = TRUE),
        beside = TRUE,
        names.arg=c("Mon","Tue","Wed","Thu","Fri"),
        col=c("blue", "red"),
        density = 30, 
        main="Week Sales")
```

Simple example of a bar plot without additional functions:

```{r}
max.temp.germany <- c(20, 28, 36, 30, 39)

barplot(max.temp.germany)
```

```{r}
# barchart with added parameters
barplot(max.temp.germany,
main = "Maximum Temperatures in Germany (July 2019)",
xlab = "Degree Celsius",
ylab = "Day",
names.arg = c("01.07.", "05.07.", "13.07", "18.07.", "20.07."),
col = "blue",
horiz = FALSE)
```

Turning to histogram, the function and some arguments are the following.

- `hist` is the function to create an histogram from the values of a vector.
- `breaks` is the important argument as it controls the width of the bars.
- `freq` determines whether the frequency (default) of the density should be plotted.

Example histogram:

```{r hist-plots, out.width="35%", fig.align="center", fig.height=7, fig.cap="Simple histograms.", fig.show="hold"}
hist(c(sales, sales2))

brks <- seq(min(s.range), max(s.range)+3, 5)
hist(c(sales, sales2),
    col=rev(heat.colors(length(brks))),
    breaks=brks,
    main="Histogram of Week Sales",
    las=1, cex.axis=0.8,
    freq=TRUE,
    xlab="Sales")

```

Basic histogram example including colors and labels:

```{r}
hist(iris$Sepal.Length, 
     breaks=30, 
     col="red", 
     xlab="Sepal Length", 
     main="Histogram Example with Colors and Labels")
```

### Other plots

Other types of plots are called by the following functions.

- `pie` for the pie (round) charts.
- `boxplot` for the box-and-whisker plot.
- `dotchart` for the Cleveland dot plot.
- ...