# R as a calculator {#rcalc}

R can also be used as a simple calculator. For instance, 25+12=`r 25+12`.

Here are a few more examples of how R can be used as a calculator with different commands:

## Usual operators

### Simple operations

The usual symbols `+, -, *, /` apply .
```{r}
2 + 6
56/ 6
```

## Order of operations

The order that arithmetic operations are performed is consistent with convention. The PEMDAS convention implies that parentheses get highest priority, followed by exponentiation, followed by multiplication and division, followed by addition and subtraction. The first example of using R in the calculator mode is given below.

```{r
2 + 9 * 4 # PEMDAS; place spaces around operators
```
[1] 38

No equal sign is necessary; pressing the return key serves that role. The response from R is [1] 38, which means that R performed the multiplication before the addition per the PEMDAS convention, giving a result of 38. The [1] that appears before the 38 indicates that the result is a vector of length one. More details concerning vectors will be given in Chapter 4. It is good programming practice to surround the operators with a single space for readabilit. This is not necessary, however, and it is perfectly acceptable to key in the R command as

```{r
> 2+9*4 # no spaces around operators; it still works
[1] 38
```

But you could imagine how difficult this would be to read if it was 30 or 40 characters long.

## Number of digits displayed

R must make a decision on how many digits to display. The default in R is to display seven digits. This can be altered, however, as will be shown subsequently.
R does not include commas to separate groups of three digits when displaying the result.

```{r
8 * 10 ˆ 3 # exponentiation first
```
[1] 8000

Now consider the calculation of a large number.

```{r
1111111 * 1111111 # no commas on constants; result in scientific notation
[1] 1.234568e+12
```

The result of this calculation is expressed in scientific notation: 1.234568 · 1012. With only seven digits displayed, it is not possible to know the exact value of the product.

By default, R displays seven digits, but more digits are stored. As an illustration, the built-in constant π is designated with pi.

```{r
pi # built in constant
```
[1] 3.141593

### Parentheses 

Parentheses are or can be used to alter the order of operations. But they are also a common source of error when they are not matched.

```{r}
(4+3)*((7-3)/(1+.05))
```
The next expression will generate an error and prevent the compilation of the whole book.
```r
(4+3)*((7-3/(1+.05))
```

Other examples of different orders through adding parantheses:

```{r
5 + (6 / 7) ˆ 2 # division first, then exponentiation, then addition
```
[1] 5.73

```{r
(2 + 1) / 5 ˆ 2 # addition first, then exponentiation, then division
```
[1] 0.12

```{r
(9 + 3 / 4) ˆ 2 # division first, then addition, then exponentiation
```
[1] 95.06

```{r
((4 + 7) / 10) ˆ 2 # addition first, then division, then exponentiation
```
[1] 1.21

### Exponents 

There are two ways of expressing the power of a number: `^` and `**`.

```{r}
3^4
3**4
```

## Unusual operators

### Special operations

The symbols `%/%` and `%%` return the entire part of the result of the division and the rest of the division, respectively.
```{r}
56/6
56%/%6
56%%6
```

## Usual functions

The common functions found in any calculator also have an equivalent on R. The following examples need no further comment.

```{r}
log(100)
sqrt(100)
```

## Unusual functions

There are several other functions applied to numbers that one does not usually find in a calculator.  
A few examples below:

```{r}
# floor / ceiling for the closest integer
floor(13.47)
ceiling(13.47)
```






