# (PART) R Basics {-}

# R as a calculator {#rcalc}

R can be used as a simple calculator. For instance, 45+17=`r 45+17` .   
Here are a few more examples of the commands.


## Usual operators

### Simple operations

The usual symbols `+, -, *, /` apply .
```{r}
2 + 6
56/ 6
```


### Parentheses 

The parentheses work as expected. But they are also a common source of error when they are not matched.

```{r}
(4+3)*((7-3)/(1+.05))
```
The next expression will generate an error and prevent the compilation of the whole book.
```r
(4+3)*((7-3/(1+.05))
```

### Exponents 

There are two ways of expressing the power of a number: `^` and `**`.

```{r}
3^4
3**4
```

## Unusual operators

### Special operations

The symbols `%/%` and `%%` return the entire part of the result of the division and the rest of the division, respectively.
```{r}
56/6
56%/%6
56%%6
```

## Usual functions

The common functions found in any calculator also have an equivalent on R. The following examples need no further comment.

```{r}
log(100)
sqrt(100)
```

## Unusual functions

There are several other functions applied to numbers that one does not usually find in a calculator.  
A few examples below:

```{r}
# floor / ceiling for the closest integer
floor(13.47)
ceiling(13.47)
```






