# Simple functions on vectors 

In this section, we illustrate calculations with a single vector.  


## Element by element evaluation

The main point to notice is how these functions are carried **element by element**. This means that the function is typically applied for each element individually and in sequence. 

The following simple example illustrates this procedure and emphasizes the effect on the length of the vectors.
```{r}
# create two numeric vector of same length
visits1 <-  c(12, 2, 45, 75, 65, 11, 3)
visits2 <- c(23, 4, 5, 78, 12, 0, 200)
length(visits1)
length(visits1) == length(visits2)

total <- visits1 + visits2
# their sum is the vector of same length, with the element by element sums

length(total)
total

```

Another clear example with the function `^2`.


```{r}
total.p2 <- total^2
total.p2

(trop <- total - 2)
length(trop)
```



## Recycling rule


With the emphasis on the length of the vectors above, one could wonder what happens when vectors' length differ. This is an important point to remember: the shortest vector has its elements **recycled** as much as it is necessary.

```{r}
l3 <- c(12, 34, 50)
l2 <- c(10, 3)
tt <- l3 + 5
tt

ltotal <- l3+l2 # recycling!!!
ltotal
```


## Functions for the whole vector

Some functions are meant to be applied to whole vector and do not necessarily return a vector of the same length of the vector to which it is applied.   
Here is an illustration of some of these functions.

```{r}
ages <- c(28, 33, 39, 56, 34, 45, 27, 40)
ages
max(ages)
sum(ages)
length(ages)

ages <- c(28, 33, 39, 56, 34, 45, 27, 40, NA)
mean(ages)

# for help on a command, simply type ? in front of it
?mean

```







