# Conditions {#conditions}

The general purpose of conditions is to control the flow of our code when executed by R.  
In R, it builds on statements such as `if` and `else`.

## `if` statement
The simplest form for a condition uses a `if` statement.  
The form is then:

```r 
if (condition) {
# code to be executed if the condition is TRUE
}
```

The key point is that R will run the code until it finds a condition that is met. If it doesn't find any, it continues to the next lines of code.  
Here is an example.



```{r}
gains <- c(10, 3, -5, 0, -4, 12, 4)
gains
sum(gains)

if (sum(gains) > 0) {
  print("Congratulations, you are winning!")
}

# change the -4 in 'gains' to -40

gains[gains==-4] <- -40
gains

if (sum(gains) > 0) {
  print("Congratulations, you are winning!")
} # notice that there is no output, because the condition was not met
```

## `else` statement

What happens when the condition is not met? It's on the user to decide. It can be nothing or... something else!

```r
if (condition) { 
# code to be executed if the condition is TRUE
} else {
# code to be executed if the condition FALSE
}
```

Here is an example.

```{r}
gains<- c(10, 3, -5, 0, -4, 12, 4)
gains[1] <- -10 # change the first element of gains
sum(gains)

if (sum(gains) > 0) {
  print("Congratulations, you are winning!")
} else {
  print("You are not winning!")
} # notice that now there is an output!
```

## `else if` statement

`else if` allows us to introduce another condition to our flow of code

```r
if (condition_1) { 
# code to be executed if the condition_1 is TRUE
} else if (condition_2) {
# code to be executed if the condition_2 is TRUE
} else {
# code to be executed if NEITHER condition_1 NOR condition_2 is TRUE
}
```

We can use as many `else if` statements as we want.  
Here is an example with only one `else if`
```{r}
gains <- c(10, 3, -5, 0, -4, 12, 4)
gains[1] <- -20 # change the first element of 'gains'
sum(gains)

if (sum(gains) > 0) {
  print("Congratulations, you are winning!")
} else if (sum(gains)==0) {
  print("You just break even!")
} else {
  print("You are losing!")
}
```

Notice the potential problem of not putting a  `else` statement at the end of the conditions system.

## `ifelse` statement

For **short** conditions, we can use `ifelse`.  
The form is
```r
ifelse(condition, value if condition met, value if value not met)
```
Here is an example.
```{r}
gains <- c(10, 3, -5, 0, -4, 12, 4)
sign <- ifelse(sum(gains)>=0,"+","-")
sign
```

## Logical operators: `&`, `|` and `!`
Logical operators are used to combine, mix or negate several conditions:
- `&` means AND
- `|` means OR
- `!` means NOT

De Morgan's laws may help here.

```{r}
(10%%2==0 & 27%%3==0) # equivalent to (TRUE and TRUE), hence TRUE
TRUE & FALSE
!TRUE 
!(TRUE & TRUE)
!(TRUE & !TRUE)
((10%%2==0 & 27%%2==0)) # equivalent to (TRUE and FALSE), hence FALSE
((10%%2==0 | 27%%2==0)) # equivalent to (TRUE or FALSE), hence TRUE
```

## Writing and interpreting a condition

Notice that what R looks for in a condition is a either a TRUE or a FALSE.  
If it encounters a TRUE, it executes the commands, otherwise, it doesn't.  
Remember there are many ways to obtain one of these two logicals. Any of these ways will work as a condition.  
Here are examples of less trivial ways of writing a condition 
```{r}
gains<- c(10, 3,-5,0,-4,12,4)

if (is.numeric(gains)) {
  print("Seems like it is a numeric vector...")
}
# here, is.numeric(gains) evaluates to TRUE, hence, the code is executed!
# the beginner's way would be to replace the condition by
# if (class(gains)=="numeric")

if (!is.factor(gains)) {
  print("Yeah! We avoided the factor...")
}
# the beginner would write if (class(gains)!="factor")
```

