# Loops (avoid them) {#loops}



## Introduction to loops with a silly example (bad R coding)

Loops allow to iterate a set of functions / lines of code over a predefined list of elements.  
They are common in many languages but are not highly regarded in R because:

  - they do not represent an efficient way of producing a result,
  - functionals, functions that return a vector such as the `apply` family, are preferred.
    
This section serves as an introduction to the functionals because it helps understand the structure of the problem.  

Now the silly example. Suppose our gains at a game over the week are, in euros,

```{r}
gains <- c(10, 3,-5,0,-4,12,4)
gains
```
Now, suppose we want to transform each gain into dollars, i.e., we must multiply each gain by 1.30. An intuitive way would be to take each gain _separately_ and multiply it 1.12. Let's do it by hand:

```{r}
gains_d <-numeric(length(gains)) # create a numeric vector of same length as gains
gains_d[1] <- gains[1]*1.12 
gains_d[2] <- gains[2]*1.12
gains_d[3] <- gains[3]*1.12
gains_d[4] <- gains[4]*1.12
gains_d[5] <- gains[5]*1.12
gains_d[6] <- gains[6]*1.12
gains_d[7] <- gains[7]*1.12
gains_d
```

We recognize that the structure of each line of code: the first line is for the element 1 of the vectors, the second line for the second element of the vectors, etc.  
Doing a loop builds on that observation. But it asks the program to make the change between each line automatically.  
In this silly case, we want the program to evaluate a line with the value 1 and then, when it has finished, do the same line but with a 2 instead of a 1. And we want it to do it **for** the elements 1, 2, 3, 4, ..., `r length(gains)`.  
This is what a loop does!


```{r}
# create an empty numeric vector of same length as gains
gains_d2 <-numeric(length(gains)) 

# now the loop
# first for i=1, then i=2, then i=3, etc... until i=length(gains)
for (i in 1:length(gains)) {  
  gains_d2[i] <- gains[i]*1.12 
}
gains_d2
```



## General form of a `for` loop

The general form of a loop is the following

```r
for (i in list-of-elements) {
# code invoking i identifier
}
```
A few words about this simple form are the following.  

  - The `i` is the identifier to be used in the code; we can use anything (e.g., `x`, `kk`, `element`, ...); we only need to be consistent and, if any, use that identifier in the code inside the loop.
  - The `in` is self-describing.
  - The list of elements is generally a vector, but it can also be a list.
  - The loop executes some code but does not need to return an object.

### No identifier in the code?

The program will run the code in the loop until it finds the identifier, replaces it with the current value of the loop and executes all the commands.  
If it doesn't find it, then the program executes everything till the end of the loop and then starts again with the next value of the loop. 
```{r}
for (kk in 1:4) {
  print("I'm working on it...")
}

for (kk in 1:5) {
  print(paste("This is loop number", kk))
}

```

## A good example of a loop

Suppose we want to know, every day, what is our cumulative gain until and including that day. This implies that we must to create another `cumulative_gains` with the same length as `gains`. 

Then, we could run the following loop.
```{r}
# create an empty vector
cumulative_gains <-numeric(length(gains)) 

cumulative_gains[1] <- gains[1]  
 
for (k in 2:length(gains)){
  cumulative_gains[k]<-cumulative_gains[k-1] + gains[k]
}
gains
cumulative_gains
```


## Looping over non-numeric elements 

The examples above could give the impression that we must loop over numeric values. But it is not so!  
The identifiers above simply took the values, one after the other, of the vector provided. Since we always gave a numeric vector, the identifier was always numeric.  
Here, we show that we can also loop over other types of vectors, for instance a character vector.  
The example is simple: we want to count the number of letters in each element of the character vector and print the result in the console. You can use the built-in function `nchar()`; `nchar("Cologne")=``r nchar("Cologne")`. 

```{r}
cities <- c("Lisbon", "Paris", "Washington", "Antananarivo")
cities

for (cit in cities) {
  print(paste(cit, "has", nchar(cit), "letters."))
}
```
To store these values in a vector `number_letters`, we could write
```{r}
number_letters <- NULL

for (cit in cities) {
  number_letters <- c(number_letters, nchar(cit)) 
}

names(number_letters) <- cities
number_letters
```



## More advanced stuff

### Loop over a `list`

We can loop over the elements of a `list`. Remember, however, that the elements of a `list` can be of any type. Therefore, we must make sure that the code inside the loop can be applied to each element of the `list`.

```{r}
mylist<- list(numbers=c(1:20), 
              cities=c("Cologne","Lisbon"), 
              results= c(T,F,F,T), 
              school=factor(c("Primary", "Secondary", "Tertiary"), ordered=TRUE))

for (kk in mylist) {
  print(length(kk))
}
```
Notice that the previous call could be done with a loop over a numeric vector.

```{r}
# (kk in 1:4)
for (kk in 1:length(mylist)) { 
  print(length(mylist[[kk]]))
}
```

### `while` loop

The command `while` introduces a loop that is run while a certain condition is satisfied. Here is an example.

```{r}
kk <- 6
while (kk < 10) {
  print(kk)
  kk <- kk + 1
}
```
Notice that the most important line is the last in the loop `kk <- kk + 1`. This makes the identifier change value, in this case, it increases of 1. 

If the identifier does not change, the program keeps running as long as the condition `(kk < 10)` is satisfied... i.e., forever! Make sure that the condition is not satisfied at some point after some loops!

```{r}
repetition <- 0
while (repetition <= 9 ) {
  print(paste("That's " , repetition, "! One more! Come on!", sep="")) 
  repetition<- repetition +1
}
```

### `break` & `next`

If we insert `break` in a loop, the program stops everything, i.e., the remaining code in the loop is skipped and the loop is not iterated over anymore.

You may want to use it in combination of a condition 
```r
if (condition==TRUE) {
  break
} 
```


```{r}
repetition <- 0
while (repetition <= 9 ) {
  if (repetition==5){
    print(paste("That's " , repetition,
                "! Well done! I know we said 10, but that's enough for today!", sep=""))
    break
  }
  print(paste("That's " , repetition, "! One more! Come on!", sep="")) 
  repetition<- repetition +1
}
```

If we insert `next` in a loop, the program skips the remainder of the code in the loop and starts with the next value in the loop. Again, it's usually combined with a condition.


