# Import data to R {#import}



The first step in order to analyze data with R is to get the data in R.  
Data out there exists in many formats and it is not always consistently recorded. In R speaking, we will need to `read` the data from a file and put it in an object, often a data frame.  
Base R has has base functions to read some types of files. For special types of files, special packages are needed.

**Important note:** these base functions are introduced here for documentation purpose only. Indeed, their task is usually better achieved by the functions developed in the `tidyverse` (see Chapter \@ref(tidyverse) and following).

## Reading rectangular data

Rectangular data is intended here as a data set with values of variables in columns and each row representing a case. Arguably, this is the simplest and probably most common way to store data sets.  
The ad hoc function in R is `read.table`. Other similar functions exist, e.g., `read.csv`, but they are essentially a wrapper of `read.table`.   
This function reads rectangular data and assigns it to an object, in almost all cases a data frame. The arguments of the function are numerous are, depending on the case, potentially very useful (check `?read.table`).  We mention here a few of them in the typical call:

```r
my.data <- read.table(...)
``` 

`file` is the name of the file to be read. It requires a valid path. It can also be an `url`, with some adaptations.  
`header` is a logical indicating whether the data's first line gives the names of the variables.  
`sep` indicates the character that separates the values between columns.

```{r}
# assuming we have the file AH001.txt in the same folder
df <- read.table(file="AH001.txt",
				header=TRUE,
				sep=",")
head(df)
```


## Read other data types

R can import virtually all types of data. For special types, a ad hoc package will be necessary. `foreign` is one of them and it allows to import data from other applications. Its functions are of the form

```r
read.XXX
```
 where the `XXX` is an extension specific to the external software. Examples include "read.dta" for Stata data files, `read.spss` for SPSS files or `read.dbf` for dBASE files.   

A very important data source type is Excel. Functions for this type of data are covered in Chapter \@ref(readr) from the part on `tidyverse`.


## Scanning a file

The base function `scan` can sometimes be useful, though it is not used to import data. `scan()` imports to a vector (or a list), which can then be used.   

- `what` gives the type of vector to be imported.  
- `skip` is the number of lines to be skipped in the file.  
- `nlines` determines the number of lines to be read and imported.

```{r}
df <- scan("AH001.txt", what=character())
#df
df <- scan("AH001.txt", what=character(), skip=1, sep=",")
#df
given.line <- scan("AH001.txt", what=character(), nlines=1, sep=",")
given.line
```





