# `tibble` {#tibble}

A tibble is the equivalent of a data frame in the `tidyverse`. The specific package of the bundle is called `tibble`.  
All the packages in the `tidyverse` work with tibbles.   
Notice that all the functions applicable to a data frame will also work on a tibble. This is because **all tibbles are data frames** (while the reverse is not true, all data frames are not necessarily tibbles).  
So why bother?  
The reasons to choose tibbles include: 

- faster calculations,
- better printing of the content,
- more information about the content,
- more useful warnings of errors,
- simpler and more limited functionalities.


## Creating a tibble

There are various ways to create a tibble.

### `as_tibble`

The simplest way is declare a data frame a tibble with the function `as_tibble`. 

```{r}
data(iris)
# print the data frame
iris

t_iris <- as_tibble(iris)
# print the tibble
class(t_iris)
t_iris
```

Another example for using as_tibble:

```{r}
data(letters)
# print the data frame
letters

t_iris <- as_tibble(letters)
# print the tibble
class(t_iris)
t_iris
```

### Manually

A tibble can also be created manually. Notice the following two points:

- vectors of length 1 (only) are recycled,  
- vectors that depend on others can be directly created.
  
```{r}
my_tbl <- tibble(
  x1 = 1:10, 
  x2 = 5, 
  x3 = x1^2 * x2
)
my_tbl
```

Another example for manually creating a tibble:

```{r}
tibble(x = 1:10, y = x ^ 3)
```


### `tribble`

A tribble is a tibble created in a transposed way. To be noticed:

- column headings are entered (as a formula) with a `~`,
- data are entered in rows,
- all values are separated by commas.

```{r}
my_tr <- tribble(
	~x1, ~x2, ~x3,
	"A", 2, 7,
	"B", 5, 12,
	"C", 4, 8 
	)
my_tr
```
















