---
title: "StringR"
output: html_document
---

Strings are not glamorous, high-profile components of R, but they do play a big role in many data cleaning and preparation tasks. The stringr package provide a cohesive set of functions designed to make working with strings as easy as possible. If you’re not familiar with strings, the best place to start is the chapter on strings in R for Data Science.

stringr is built on top of stringi, which uses the ICU C library to provide fast, correct implementations of common string manipulations. stringr focusses on the most important and commonly used string manipulation functions whereas stringi provides a comprehensive set covering almost anything you can imagine. If you find that stringr is missing a function that you need, try looking in stringi. Both packages share similar conventions, so once you’ve mastered stringr, you should find stringi similarly easy to use.

Installation
# Install the released version from CRAN:
install.packages("stringr")

# Install the cutting edge development version from GitHub:
# install.packages("devtools")
devtools::install_github("tidyverse/stringr")
Cheatsheet


Usage
All functions in stringr start with str_ and take a vector of strings as the first argument.

```{r}
x <- c("stadium", "football", "players", "halftime", "referee", "goalkeeper")
str_length(x) 

str_c(x, collapse = ", ")

str_sub(x, 1, 3)
```

There are seven main verbs that work with patterns:

str_detect(x, pattern) tells you if there’s any match to the pattern.

```{r}
str_detect(x, "[a]")
```

str_count(x, pattern) counts the number of patterns.

```{r}
str_count(x, "[a]")
```

str_subset(x, pattern) extracts the matching components.

```{r}
str_subset(x, "[r]")
```

str_locate(x, pattern) gives the position of the match.

```{r}
str_locate(x, "[a]")
```

str_extract(x, pattern) extracts the text of the match.

```{r}
str_extract(x, "[a]")
```

str_match(x, pattern) extracts parts of the match defined by parentheses.

# extract the characters on either side of the vowel
```{r}
str_match(x, "(.)[ae](.)")
```

str_replace(x, pattern, replacement) replaces the matches with new text.

```{r}
str_replace(x, "[aei]", "X")
```

str_split(x, pattern) splits up a string into multiple pieces.
```{r}
str_split(c("a,b", "c,d,e"), ",")
```

