--- 
title: "An Introduction to R for the Field of Data Science"
author: "Moritz Ebach (400104774), Niklas Strolz (400098595), Frederik Miebach (KOWH116385)"
date: "Last update: `r format(Sys.Date(),'%B %d, %Y')`"
site: bookdown::bookdown_site
output: bookdown::gitbook
documentclass: report
bibliography: [book.bib, packages.bib, online.bib]
biblio-style: apalike
link-citations: yes
github-repo: rstudio/bookdown-demo
description: "These are notes gathering our experience to set up R-Studio and use R for data science."
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      cache = TRUE,
                      collapse= TRUE,
                      comment = "#R>")

library(tufte)
library(tidyverse)
library(magrittr)
library(gridExtra)
library(readxl)
library(parallel)
library(hflights)
```



```{r include=FALSE, eval=FALSE}
# this generates an error message in my machine, hence the eval=FALSE
# automatically create a bib database for R packages
knitr::write_bib(c(
  .packages(), 'bookdown', 'knitr', 'rmarkdown', 'tidyverse', 'tufte', 'gridExtra', 'magrittr', 'readxl', 'parallel'
), 'packages.bib')

```


# Foreword {-}



> "The statistic is the rhetoric of the empirical sciences."   
>
> --- paraphrase of Alexander Eiler's quote on statistics 



Students with a focus on economics and science have to work a lot with statistical data sets and calculations in order to be able to carry out evaluations. Traditionally, educational institutions work with formula collections and calculators, which has several disadvantages. On the one hand, solutions have to be documented in a laborious way, on the other hand, statistical calculations with formula collections are very time-consuming. The programming language R in combination with R Studio enables students like us to analyze and visualize complex data sets and to adapt them according to our own criteria with less effort compared to the traditional methods. With this book we want to give a practical guide on how to apply these benefits of R to beginners.

So we focus on three questions.

1. What steps must be undertaken to be able to work productively with R and R studio?
2. How can R be used for data science using practical examples?
3. What does a complex project look like in R and what steps must be taken to solve it?