--- 
title: "Learning R for Data Science"
author: "Antonio Fidalgo"
date: "Last update: `r format(Sys.Date(),'%B %d, %Y')`"
site: bookdown::bookdown_site
output: bookdown::gitbook
documentclass: book
bibliography: [book.bib, packages.bib, online.bib]
biblio-style: apalike
link-citations: yes
github-repo: rstudio/bookdown-demo
description: "These are notes gathering my experience in R for data science."
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      cache = TRUE,
                      collapse= TRUE,
                      comment = "#R>")

library(tufte)
library(tidyverse)
library(magrittr)
library(gridExtra)
library(readxl)
library(parallel)
library(hflights)
```



```{r include=FALSE, eval=FALSE}
# this generates an error message in my machine, hence the eval=FALSE
# automatically create a bib database for R packages
knitr::write_bib(c(
  .packages(), 'bookdown', 'knitr', 'rmarkdown', 'tidyverse', 'tufte', 'gridExtra', 'magrittr', 'readxl', 'parallel'
), 'packages.bib')

```


# Foreword {-}



> "Data science is like teenage sex: everyone talks about it, nobody really knows how to do it, everyone thinks everyone else is doing it, so everyone claims they are doing it..."   
>
> --- paraphrase of Dan Ariely's quote on big data 



There is already a large number of excellent and free references for learning data science in R. The list would be too vast, but two names stand out: [Hadley Wickham](http://hadley.nz/) and [Yihui Xie](https://yihui.name/en/about/).    
Work of the first includes the books @wickham2016 and @wickham2014 (freely available, including source code, on [github.com/hadley](https://github.com/hadley)) as well as some of the most popular packages in R such as  `ggplot2` [@R-ggplot2] and `tidyverse`[@R-tidyverse].  
Work of the second is used precisely in the current file thanks to the `rmarkdown` package [@R-rmarkdown], `knitr` [@R-knitr] and `bookdown` [@R-bookdown]. His source code is also on [github.com/yihui/](https://github.com/yihui/).  

So, the question arises of why one should even bother to write on the topic. The present "book" is justified on the following grounds.  

1. There is no better way of learning data science and R than **doing oneself** the pages of code.  
2. The material gathered here is a **personal selection** made on what I judge most relevant for my work, the most important techniques in general or, sometimes, the least obvious for the regular practitioner.  
3. These notes serve as an **archive** of what I did in the domain so that I can easily access it in the future.







